---
title: "day1_clusterProfiler_ora"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---


```{r}
load("demo_ora.RData")
```

```{r, message=FALSE}
library(clusterProfiler)
```

**You need to make sure the gene IDs are Entrez IDs.** The following function
helps to automatically convert gene IDs to Entrez IDs. The input can be a vector
of genes or a gene expression matrix.

```{r}
convert_to_entrez_id = function(x) {
	if(is.matrix(x)) {
		map = cola:::guess_id_mapping(rownames(x))
		if(is.function(map)) {
		    new_rn = map(rownames(x))
		} else {
		    new_rn = map[rownames(x)]
		}
		l = is.na(new_rn)

		x = x[!l, , drop = FALSE]
		new_rn = new_rn[!l]

		x2 = do.call(rbind, tapply(1:nrow(x), new_rn, function(ind) {
			colMeans(x[ind, , drop = FALSE])
		}))
		return(x2)

	} else if(is.numeric(x)) {
		map = cola:::guess_id_mapping(names(x))
		x2 = s
		if(is.function(map)) {
		    names(x2) = map(names(x))
		} else {
		    names(x2) = map[names(x)]
		}
		x2 = x2[!is.na(names(x2))]
		x2 = tapply(x2, names(x2), mean)
		return(x2)
	} else {
	    map = cola:::guess_id_mapping(x)
	    if(is.function(map)) {
	        x2 = map(x)
	    } else {
	        x2 = map[x]
	    }
	    x2 = x2[!is.na(x2)]
	    x2 = unique(x2)
	    return(x2)
	}
}
```


We convert the diff genes `diff_gene` to Entrez IDs. Note some genes are lost 
due to the conversion.

```{r}
head(diff_gene)
diff_gene = convert_to_entrez_id(diff_gene)
head(diff_gene)
length(diff_gene)
```

Next we perform ORA on different gene sets

1. GO enrichment

```{r, message = FALSE}
library(org.Hs.eg.db)
tb = enrichGO(gene = diff_gene, ont = "BP", OrgDb = org.Hs.eg.db)
head(tb)
```

2. KEGG enrichment


```{r, message = FALSE}
tb = enrichKEGG(gene = diff_gene, organism = "hsa")
head(tb)
```
3. Reactome enrichment

```{r}
library(ReactomePA)
tb = enrichPathway(gene = diff_gene)
head(tb)
```
4. DO enrichment

```{r}
library(DOSE)
tb = enrichDO(gene = diff_gene, ont = "DO")
head(tb)
```

5. MSigDB enrichment

There is no built-in function specific for MSigDB gene sets, but there is a universal function `enrichr()` which accepts
manually-specified gene sets. The gene sets object is simply a two-column data frame:

- the first column is the gene set ID
- the second column is the gene ID

```{r}
library(msigdbr)
gene_sets = msigdbr(category = "H")
map = gene_sets[, c("gs_name", "entrez_gene")]

tb = enricher(gene = diff_gene, TERM2GENE = map)
head(tb)
```

