---
title: "Get pathways from Reactome"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---


Reactome is another popular pathway database. It organise pathways in an hierarchical category, which contains pathways and sub pathways
or pathway components. The up-to-date pathway data can be direclty found at https://reactome.org/download-data.

There is a **reactome.db** on Bioconductor. Similar as other annotation packages. Users can type `reactome.db::` with two continuous tabs to see the objects
supported in the package. In it, the important objects are 

- `reactomePATHID2EXTID` contains mappings between reacotme pathway IDs and gene entrez IDs
- `reactomePATHID2NAME` contains pathway names

```{r}
library(reactome.db)
tb = toTable(reactomePATHID2EXTID)
head(tb)
p2n = toTable(reactomePATHID2NAME)
head(p2n)
```

In the previous code, we use the function `toTable()` to retrieve the data as data frames. Readers may try `as.list()` on the two objects and compare the output.

Reactome also contains pathway for multiple organisms. In the reactome ID, teh second section contains the organism, e.g. in previous output `HSA`. 

```{r}
table( gsub("^R-(\\w+)-\\d+$", "\\1", p2n[, 1]) )
```

Again, **reactome.db** only contains pathways as list of genes. 
