---
title: "Getting gene sets from BioMart"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---


The **org.\*.db** provided by Bioconductor should be enough for most of the analysis tasks. However, there may be users
who mainly work on non-model organism or microbomens where there is no such pre-compiled package on Bioconductor. 
In this case, they can use the **biomaRt** package to get GO gene sets for the non-common organisms.


BioMart (https://www.ensembl.org/info/data/biomart) is a web-based integrated sevice from Ensembl database which can be used to extract information from a huge number of external databases. 
The companion R package **biomaRt** provides a programatical interface to directly access BioMart purely in R. Using **biomaRt** to extract GO gene sets is a little bit complex
and I will demonstrate it step by step.

1. Connect to a mart (database). 

```{r}
library(biomaRt)
ensembl = useEnsembl(biomart = "genes")
```

By default, it uses the newest "core" Ensembl database. To use other organisms, readers
can switch to another mart. All supported marts can be obtained with the function `listEnsemblGenomes()`.

```{r}
listEnsemblGenomes()
```

We only need to use the mart with "\_mart" suffix. For example, to use the fungi mart, we can select
the mart by:

```{r, eval = FALSE}
ensembl_fungi = useEnsemblGenomes("fungi_mart")
```


2. After a specific mart is selected, we need to select a dataset for a
   specific organism. To find a valid value of dataset,
   we use `listDatasets(ensembl)`. 

```{r}
datasets = listDatasets(ensembl)
dim(datasets)
head(datasets)
```

`datasets` contains names of datasets for all organisms supported in a given mart, which is normally very long. We can use regular expression to match the organism. For example, we want to select the dataset for giant panda.

```{r}
ind = grep("panda", datasets$description, ignore.case = TRUE)
ind
datasets[ind, ]
```

The name of the dataset is in the first column of `datasets`. Here for giant panda, the corresponding dataset is `"amelanoleuca_gene_ensembl"`. We update the `ensembl` object by
selecting the dataset.

```{r}
dataset = "amelanoleuca_gene_ensembl"
ensembl = useDataset(dataset = dataset, mart = ensembl)
```

3. Get the associations between genes and GO terms. The dataset can be thought as a giant table, and we only need to extract the columns of gene and GO IDs. The core function `getBM()` is to extract the corresponding "tables" which is called "attributes". Now again, we need to know what are the valid attributes for genes and GO terms. 

The function `listAttributes(ensembl)` returns all supported attributes (or columns) in a dataset. 

```{r}
attr = listAttributes(ensembl)
dim(attr)
head(attr)
```

As it shows, `attr` contains `r nrow(attr)` rows. It is quite handy to find
the correct attributes for genes and GO terms. Normally we need to try to
match them with different regular expressions. For gene ID, it is relatively simple because 
the first row in `attr` contains the attribute name `"ensembl_gene_id"`

To find GO-related attributes, we can try the following regular expression:

```{r}
ind = grep("GO", attr$description)
ind
attr[ind, ]
```

Now we can see the attribute for GO IDs are `"go_id"` and for GO namespaces is `"namespace_1003"`.
We select the three attributes `c("ensembl_gene_id", "go_id", "namespace_1003")` and retrieve 
the data with `getBM()`.

```{r}
tb_go = getBM(attributes = c("ensembl_gene_id", "go_id", "namespace_1003"), 
    mart = ensembl)
head(tb_go)
```

4. To make it simple, we only use GO terms in BP namespace and convert it to a list.

```{r}
tb = tb_go[tb_go$namespace_1003 == "biological_process", , drop = FALSE]
gs = split(tb$ensembl_gene_id, tb$go_id)
length(gs)
```

5. Unfortunately, we can not directly use `gs` because genes associated to a child term have
not been merged to the parent terms. Thus, gene sets in `gs` is not complete. We need to use 
`GOBPOFFSPRING` to find all offspring terms of a GO term.

In the following code, we use `keys()` to extract all GO terms that have offsprings. It is the same 
as using `names(as.list(GOBPOFFSPRING))` and `unique(toTable(GOBPOFFSPRING)[, 2])`. For each GO term, we take the union of the GO term and its offspring (note `GOBPOFFSPRING[[nm]]` returns a vector of GO terms which are the offspring of `nm`), and finally take all the genes annotated to them.

```{r, eval = FALSE}
bp_terms = keys(GOBPOFFSPRING)
gs2 = lapply(bp_terms, function(nm) {
    go_id = c(nm, GOBPOFFSPRING[[nm]])
    unique(unlist(gs[go_id]))
})
names(gs2) = bp_terms
gs2 = gs2[sapply(gs2, length) > 0]
```

#### BioMartGOGeneSets

To obtain the gene sets, use the function `getBioMartGOGeneSets()`. Also you need to provide
the "dataset". Here we use a different dataset: `"mmusculus_gene_ensembl"` (mouse).

```{r}
library(BioMartGOGeneSets)
lt = getBioMartGOGeneSets("mmusculus_gene_ensembl")
length(lt)
lt[1]
```

The variable `lt` is a list of vectors where each vector corresponds to a GO gene set with Ensembl
IDs as gene identifiers.

You can try the following command and see what will be printed:

```{r, eval = FALSE}
lt = getBioMartGOGeneSets("mouse")
```

In `getBioMartGOGeneSets()`, argument `as_table` can be set to `TRUE`, then the function returns
a data frame.

```{r}
tb = getBioMartGOGeneSets("mmusculus_gene_ensembl", as_table = TRUE)
head(tb)
```

Argument `ontology` controls which category of GO gene sets. Possible values should be `"BP"`, `"CC"`
and `"MF"`.

```{r, eval = FALSE}
getBioMartGOGeneSets("mmusculus_gene_ensembl", ontology = "BP") # the default one
getBioMartGOGeneSets("mmusculus_gene_ensembl", ontology = "CC")
getBioMartGOGeneSets("mmusculus_gene_ensembl", ontology = "MF")
```

Last, argument `gene_id_type` can be set to `"entrez_gene"` or `"gene_symbol"`, then genes in the gene sets
are in Entrez IDs or gene symbols. Note this depends on specific organisms, that not every organism supports 
Entrez IDs or gene symbols.

```{r}
lt = getBioMartGOGeneSets("mmusculus_gene_ensembl", gene_id_type = "entrez_gene")
lt[1]
```