---
title: "Getting gene sets from MSigDB"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---

Molecular signature database is a mauanlly curated gene set databases. Initially, it was proposed as a comparion dataset for the original GSEA paper. Later it has been separated out and developed independently. In the first version in 2005, there were only xx gene set categories and in total xx gene sets. Now in the newest version of MisgDb, there are xx gene sets and in total xx gene sets. It provides gene sets focusing on various aspects of xxx.

The MsigDb website already provide gene sets as text-based files, which is very easy to parse. Taking the original c2 gene sets catague as an example. The gene sets are stored in a file in `.gmt` format. `gmt` format is as follows:

```
gene_set_1  gene set description    gene1   gene2   gene3
gene_set_2  gene set description    gene4   gene5
...
```


To read the gmt file

```{r}
ln = readLines("c2.symbols.gmt")
ln = strsplit(ln, "\t")
gs = lapply(ln, function(x) x[-(1:2)])
names(gs) = sapply(ln, function(x) x[1])
```

There is already an R package **msigdbr**.

```{r, message = FALSE}
library(msigdbr)
```

Which species are supported. Please note MSigDB only provides gene sets for human,
while **msigdbr** supports more species by annotating the homologous genes.

```{r}
msigdbr_species()
```

To obtain all gene sets:

```{r}
all_gene_sets = msigdbr() # by default it is human
dim(all_gene_sets)
```

```{r}
head(all_gene_sets)
```

```{r}
as.data.frame(head(all_gene_sets))
```

All categories of gene sets:

```{r}
as.data.frame(msigdbr_collections())
```

E.g., we want to extract genesets in C2 category and CP:KEGG sub-category:

```{r}
gene_sets = msigdbr(category = "C2", subcategory = "CP:KEGG")
gene_sets
```
