---
title: "Get pathways from KEGG"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---

Kyoto Encyclopedia of Genes and Genomes (KEGG) is a comprehensive database of
genomic and molecular data for variaty of organisms. Its sub-database the
pathway database is a widely used gene set database used in current studies.
In KEGG, pathways are manually curated and number of genes in pathways are
intermediate

Keep in mind, because KEGG cann to support its maintainness, KEGG pathways are only free for academic users. 

### KEGG API


KEGG provides its data via a REST API (https://rest.kegg.jp/). There are
several commands that can be used to retrieve specific types of data. The URL form of the request is 

```
https://rest.kegg.jp/<operation>/<argument>[/<argument2[/<argument3> ...]]
```

To get the KEGG pathway gene sets, we will use the operator `link` and `list`.

The `link` operator returns the mapping between two sources of information. We can use
the following command to get the mappings between genes and pathways for human.


```{r}
df1 = read.table(url("https://rest.kegg.jp/link/pathway/hsa"), 
    sep = "\t")
head(df1)
```

In the example, `url()` construct a connection object that directly transfer
data from the remote URL. In output, the first column contains Entrez ID
(users may remove the `"hsa:"` prefix for downstream analysis) and the second
columncontains KEGG pathways IDs (users may remove the `"path:"` previx).

To get the full name of pathways, use the `list` command:

```{r}
df2 = read.table(url("https://rest.kegg.jp/list/pathway/hsa"), 
    sep = "\t")
head(df2)
```


### Packages 

#### KEGGREST

There are two Bioconductor packages for retrieving pathway data from KEGG. Both of them are based on KEGG REST API. The first one is the package **KEGGREST** which
implements a full interface to access KEGG data in R. All the API from KEGG REST service are suppoted in **KEGGREST**. For example, to get the mapping
between genes and pathways, the function `keggLink()` can be used.

```{r}
library(KEGGREST)
pathway2gene = keggLink("pathway", "hsa")
head(pathway2gene)
```

The returned object `pathway2gene` is a named vector, where the names corresponding to the source and the values correspond to the target. Readers can try to
execute `keggLink("hsa", "pathway")` to compare the results.

The named vectors are not common for downstream gene set analysis. A more used format is a data frame. We can simply converted them as:

```{r}
p2g_df = data.frame(gene_id = gsub("hsa:", "", names(pathway2gene)),
    pathway_id = gsub("path:", "", pathway2gene))
head(p2g_df)
```

In the pathway ID, the prefix in letters corresponds to the organism, e.g. `hsa` for human. 
To get a list of pathway names:

```{r}
head(keggList("pathway", "hsa"))
```

Last but not the least, another useful function in **KEGGREST** is `keggGet()` which implements the `get` command from REST API. With this function users can download
images and KGML of pathways.

```{r}
img = keggGet("hsa05310", "image")
kkgml = keggGet("hsa05310", "kgml")
``` 

However, the conf file is not supported by `keggGet()`. Users need to directly read from the URL

```{r}
conf = read.table(url("https://rest.kegg.jp/get/hsa05310/conf"),
    sep = "\t")
```

The conf file contains coordinate of genes or nodes in the image. It is useful if users want to highlight genes in the image.

```{r, echo = FALSE}
img_width = dim(img)[2]
img_height = dim(img)[1]
```

For example, we want to highlight the ninth row in `conf`

```{r}
conf[9, ]
```

This is a rectangle with two diagnols at `[142, 287]` and `[188, 304]`. With these coordinates,
we can find the position in the image and later add a rectangle to it.

```{r, fig.width = 8, fig.height = img_height/img_width*8, out.width = "100%"}
library(grid)

img_width = dim(img)[2]
img_height = dim(img)[1]
pushViewport(viewport(xscale = c(1, img_width), yscale = c(1, img_height),
    width = unit(img_width/img_height, "snpc"), height = unit(1, "snpc")))
grid.raster(img)
grid.rect(x = 142+1, y = img_height - 304, width = 188-142, height = 304-287, 
    default.units = "native", just = c("left", "bottom"),
    gp = gpar(col = "red", fill = NA, lwd = 2))
popViewport()
```

Here note since the coordinate starts from teh top left corner of the plot, while in the plotting 
coordinate, it starts from bottom left.

#### clusterProfiler

The second Bioconductor pacakge **clusterProfiler** has a simple function `download_KEGG()` which accepts the prefix of a organism and returns a list of two data frames,
one for the mapping between genes and pathways and the other for the full name of pathways.

```{r}
lt = clusterProfiler::download_KEGG("hsa")
head(lt$KEGGPATHID2EXTID)
head(lt$KEGGPATHID2NAME)
```

The two packages mentioned above do not provide data for teh network representation of pathways. In Chapter x I will demonstrate how to read and process pathways as networks. Here
we simply treat pathways as lists of genes and we ignore the relations of genes .
