---
title: "day1_ora_example"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---


# Examples of ORA analysis

```{r}
load("demo_ora.RData")
```

The lengths of the gene lists:

```{r}
length(bg_gene)
length(diff_gene)
length(gene_set)
length(intersect(diff_gene, gene_set))
```

We fill into the 2x2 contigency table:

|      | in the set | not in the set| total |
|------|------------:|---------------:|-------:|
|DE    |  $n_{11}$    |     $n_{12}$    |  $n_{1+}$ |
|not DE|  $n_{21}$    |     $n_{22}$    |  $n_{2+}$ |
|total |  $n_{+1}$    |     $n_{+2}$    |  $n$ |

<style>
.table {
width:400px;
}
</style>

Then we fill all values into the 2x2 table:

|      | in the set | not in the set| total |
|------|------------:|---------------:|-------:|
|DE    |  14    |     954   |  968 |
|not DE|  186    |     37438   |  37624 |
|total |  200    |     38392   |  38592 |


**By hypergeometric distribution**

All genes are separated into two distinct sets: in the gene set and not in the gene set:

```{r}
1 - phyper(14-1, 200, 38392, 968)
```

The separation can be performed in the other dimension: diff genes and non-diff genes:

```{r}
1 - phyper(14-1, 968, 37624, 200)
```

**By Fisher's Exact test**

```{r}
fisher.test(matrix(c(14, 186, 954, 37438), nrow = 2))
```

**By Binomial distribution**

p = probability of a gene being diff:

```{r}
1 - pbinom(14-1, 200, 968/38592)
```

Or p = probability of a gene being in the set:

```{r}
1 - pbinom(14-1, 968, 200/38592)
```

**By Chi-square test**

```{r}
chisq.test(matrix(c(14, 186, 954, 37438), nrow = 2), correct = FALSE)
```

**By two-sample _z_-test**

```{r}
p1 = 14/200
p2 = 954/38392
p = 968/38592

z = abs(p1 - p2)/sqrt(p*(1-p))/sqrt(1/200 + 1/38392)
2*pnorm(z, lower.tail = FALSE)
z^2
```

We can also do in the other dimension, the tests are identical:

```{r}
p1 = 14/968
p2 = 186/37624
p = 200/38592

z = abs(p1 - p2)/sqrt(p*(1-p))/sqrt(1/968 + 1/37624)
2*pnorm(z, lower.tail = FALSE)
z^2
```

**Test the speed of various tests:**

```{r}
library(microbenchmark)

microbenchmark(
	hyper = 1 - phyper(13, 200, 38392, 968),
	fisher = fisher.test(matrix(c(14, 186, 954, 37438), nrow = 2)),
	binom = 1 - pbinom(13, 968, 200/38592),
	chisq = chisq.test(matrix(c(14, 186, 954, 37438), nrow = 2), correct = FALSE),
	ztest = {
		p1 = 14/200
		p2 = 954/38392
		p = 968/38592

		z = abs(p1 - p2)/sqrt(p*(1-p))/sqrt(1/200 + 1/38392)
		2*pnorm(z, lower.tail = FALSE)
	},
	times = 1000
)
```

**The effect of background size**

```{r}
1 - phyper(13, 200, 38392, 968)
1 - phyper(13, 200, 38392 - 10000, 968)
1 - phyper(13, 200, 38392 - 20000, 968)
1 - phyper(13, 200, 38392 + 10000, 968)
1 - phyper(13, 200, 38392 + 20000, 968)
```

**Sensitivity of the small number of diff genes in the set**

```{r}
1 - phyper(13, 200, 38392, 968)
1 - phyper(13-3, 200, 38392, 968)
1 - phyper(13+3, 200, 38392, 968)
```
