---
title: "day1_gsea_clusterProfiler"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---

**Note in clusterProfiler, p-values are calculated by gene permutation!**

```{r, message = FALSE}
load("gene_diff_score.RData")
gene_diff_score = sort(gene_diff_score, decreasing = TRUE)
head(gene_diff_score)
```

```{r, message = FALSE}
library(clusterProfiler)
library(org.Hs.eg.db)
res = gseGO(geneList=gene_diff_score, OrgDb = org.Hs.eg.db, ont = "BP")
res
```

`gseaplot()` draws the GSEA plot:


```{r}
gseaplot(res, geneSetID=1)
```

If using the MsigDB gene sets, similar as in `day1_clusterProfiler_ora.html`, we use the "H" gene sets (hallmark gene sets).

```{r}
library(msigdbr)
gene_sets = msigdbr(category = "H")
map = gene_sets[, c("gs_name", "entrez_gene")]
```

This time, we use the function `GSEA()`:

```{r, message = FALSE}
tb = GSEA(geneList = gene_diff_score, TERM2GENE = map)
head(tb)
```

More usages of **clusterProfiler**, please go to https://yulab-smu.top/biomedical-knowledge-mining-book/index.html.


## For other organism


We take pig as an example.

```{r}
library(org.Ss.eg.db)
gene = keys(org.Ss.eg.db)
gene_diff_score = rnorm(length(gene))
names(gene_diff_score) = gene
gene_diff_score = sort(gene_diff_score, decreasing = TRUE)
res = gseGO(geneList=gene_diff_score, OrgDb = org.Ss.eg.db, ont = "BP", pvalueCutoff = 1)
res
```

Or use other gene sets

```{r}
gene_sets = msigdbr(species = "pig", category = "H")
map = gene_sets[, c("gs_name", "entrez_gene")]
tb = GSEA(geneList = gene_diff_score, TERM2GENE = map, pvalueCutoff = 1)
```


Or use a "rare" organism:


```{r}
library(BioMartGOGeneSets)
# randomly select 1000 genes
gene = getBioMartGenes("cclementina_eg_gene")$ensembl_gene_id
gene_diff_score = rnorm(length(gene))
names(gene_diff_score) = gene
gene_diff_score = sort(gene_diff_score, decreasing = TRUE)

map = getBioMartGOGeneSets("cclementina_eg_gene", as_table = TRUE)

tb = GSEA(geneList = gene_diff_score, TERM2GENE = map, pvalueCutoff = 1)
```