---
title: "day2_EnrichmentBrowser"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE)
```

Package **EnrichmentBrowser** integrates a lot of GSEA methods. The integrated methods are:

```{r}
library(EnrichmentBrowser)
sbeaMethods()
```

**EnrichmentBrowser** needs a special format (in `SummarizedExperiment`) as input.
Condition labels should be stored in a column "GROUP". Log2 fold change and adjusted p-values should be saved in "FC" and "ADJ.PVAL" columns.


```{r}
library(CePa)
condition = read.cls("P53.cls", treatment = "MUT", control = "WT")$label
expr = read.gct("P53_collapsed_symbols.gct")


library(SummarizedExperiment)
se = SummarizedExperiment(assays = SimpleList(expr = expr))
colData(se) = DataFrame(GROUP = ifelse(condition == "WT", 1, 0))

l = condition == "WT"

library(genefilter)
tdf = rowttests(expr, factor(condition))
rowData(se) = DataFrame(FC = log2(rowMeans(expr[, l])/rowMeans(expr[, !l])),
                        ADJ.PVAL = p.adjust(tdf$p.value))
se
```

**Note, to run `eaBrowse()`, you need to explicitly convert to ENTREZID.**

```{r}
se = idMap(se, org = "hsa", from = "SYMBOL", to = "ENTREZID")  # !! Gene ID must be converted to EntrezID
```

We load the hallmark gene sets by package **msigdbr**.

**When using the Entrez ID, make sure the "numbers" are converted to "characters".**

```{r}
library(msigdbr)
gs = msigdbr(category = "H")
gs = split(gs$human_entrez_gene, gs$gs_name)  # Entrez ID must be used
gs = lapply(gs, as.character)                 # be careful Entrez ID might be wrongly used as integers, convert them into characters
```

Simply call `sbea()` function with a specific method:

**Note now you can use `eaBowse(res)` to create the tiny website for detailed results.**

```{r, mess}
res = sbea(method = "gsea", se = se, gs = gs)
tb = gsRanking(res, signif.only = FALSE)
```


Next we run all supported GSEA methods in **EnrichmentBrowser**.

```{r}
all_gsea_methods = sbeaMethods()
all_gsea_methods
```



```{r, eval = !file.exists("enrichmentbrowser_res_list.RData"), results="none"}
all_gsea_methods = setdiff(all_gsea_methods, "padog")
res_list = lapply(all_gsea_methods, function(method) {
    sbea(method = method, se = se, gs = gs)
})
names(res_list) = all_gsea_methods
```

```{r, echo = FALSE, eval = !file.exists("enrichmentbrowser_res_list.RData")}
save(res_list, file = "enrichmentbrowser_res_list.RData")
```


```{r, echo = FALSE}
load("enrichmentbrowser_res_list.RData")
```

We compare the significant gene sets from different methods.

```{r}
tb_list = lapply(res_list, gsRanking)
tb_list
```

```{r}
tb_list = tb_list[sapply(tb_list, length) > 0]

library(ComplexHeatmap)
cm = make_comb_mat(lapply(tb_list, function(x) x[[1]]))
UpSet(cm,
    top_annotation = upset_top_annotation(cm, add_numbers = TRUE),
    right_annotation = upset_right_annotation(cm, add_numbers = TRUE)
)
```
