---
title: "day3_GSVA"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---

We need an expression matrix as well as a collection of gene sets.

```{r}
library(CePa)
condition = read.cls("P53.cls", treatment = "MUT", control = "WT")$label
expr = read.gct("P53_collapsed_symbols.gct")

ln = strsplit(readLines("c2.symbols.gmt"), "\t")
gs = lapply(ln, function(x) x[-(1:2)])
names(gs) = sapply(ln, function(x) x[1])
```

Running **GSVA** analysis is simple. But note the gene IDs in `expr` should match the gene IDs in `gs`.

The returned value `gs_mat` is a set-sample matrix.

```{r}
library(GSVA)
gs_mat = gsva(expr, gs, verbose = FALSE)
dim(gs_mat)
```

One good thing is you can visualize the set-level variations:


```{r, message = FALSE}
library(ComplexHeatmap)
Heatmap(gs_mat, top_annotation = HeatmapAnnotation(cond = condition),
  column_split = condition)
```

`gs_mat` can be used for downstream analysis.

E.g. we can apply t-test on each row of `gs_mat` to test whether the geneset-level
profile has difference between the two conditions.

```{r, message=FALSE}
library(genefilter)
tdf = rowttests(gs_mat, factor(condition))
tdf$fdr = p.adjust(tdf$p.value, "BH")
```

How many gene sets are significant?

```{r}
tdf[tdf$fdr < 0.05, ]
```

As a comparison, we also perform a normal GSEA analysis. We use the t-value as the gene-level score:

```{r}
s = rowttests(expr, factor(condition))[, "statistic"]
names(s) = rownames(expr)
```

Here we also test to the c2 gene set collection, so we need to convert `gs` to the format **clusterProfiler** accepts:

```{r, message = FALSE}
map = data.frame(
    gene_set = rep(names(gs), times = sapply(gs, length)),
    gene = unlist(gs)
)
    
library(clusterProfiler)
tb = GSEA(geneList = sort(s, decreasing = TRUE), TERM2GENE = map)
head(tb)
```


Actually comparing results from GSVA and GSEA, it is obvious that GSEA is better.


