---
title: "day3_topology_based_pathway_enrichment"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE)
```

We read the expression matrix as well as the condition labels.

```{r}
library(CePa)
condition = read.cls("P53.cls", treatment = "MUT", control = "WT")$label
expr = read.gct("P53_collapsed_symbols.gct")
```

We convert the gene IDs to Entrez ID because later the genes in pathways are all in Entrez ID.

```{r}
convert_to_entrez_id = function(x) {
	if(is.matrix(x)) {
		map = cola:::guess_id_mapping(rownames(x))
		if(is.function(map)) {
		    new_rn = map(rownames(x))
		} else {
		    new_rn = map[rownames(x)]
		}
		l = is.na(new_rn)

		x = x[!l, , drop = FALSE]
		new_rn = new_rn[!l]

		x2 = do.call(rbind, tapply(1:nrow(x), new_rn, function(ind) {
			colMeans(x[ind, , drop = FALSE])
		}))
		return(x2)

	} else if(is.numeric(x)) {
		map = cola:::guess_id_mapping(names(x))
		x2 = s
		if(is.function(map)) {
		    names(x2) = map(names(x))
		} else {
		    names(x2) = map[names(x)]
		}
		x2 = x2[!is.na(names(x2))]
		x2 = tapply(x2, names(x2), mean)
		return(x2)
	} else {
	    map = cola:::guess_id_mapping(x)
	    if(is.function(map)) {
	        x2 = map(x)
	    } else {
	        x2 = map[x]
	    }
	    x2 = x2[!is.na(x2)]
	    x2 = unique(x2)
	    return(x2)
	}
}
```

```{r}
expr = convert_to_entrez_id(expr)
```

Package **EnrichmentBrowser** supports variety of topology-based pathway enrichment methods.
It needs a special format (in `SummarizedExperiment`) as input.

Condition labels should be stored in a column "GROUP":

```{r}
library(SummarizedExperiment)
se = SummarizedExperiment(assays = SimpleList(expr = expr))
colData(se) = DataFrame(GROUP = ifelse(condition == "WT", 1, 0))
```

Log2 fold change and adjusted p-values should be saved in "FC" and "ADJ.PVAL" columns:

```{r}
l = condition == "WT"

library(genefilter)
tdf = rowttests(expr, factor(condition))
rowData(se) = DataFrame(FC = log2(rowMeans(expr[, l])/rowMeans(expr[, !l])),
                        ADJ.PVAL = p.adjust(tdf$p.value, "BH"))
```


Now we have `se` set up:

```{r}
se
```

Get KEGG pathways.

```{r}
library(EnrichmentBrowser)
hsa.gs = getGenesets(org = "hsa", db = "kegg")
hsa.grn = compileGRN(org="hsa", db="kegg", map2entrez=FALSE)
```

We put expression data and pathway data to `nbea()` function. Here we use `ggea` method.

```{r}
res = nbea(method = "ggea", se = se, gs = hsa.gs, grn = hsa.grn)
gsRanking(res)
```

`eaBrowse()` visualizes the results in a Shiny app:

```{r, eval=FALSE}
eaBrowse(res)
```

All topology-based methods:

```{r}
all_tp_methods = nbeaMethods()
all_tp_methods
```

We run all methods. Note some methods are extremely slow to run. `ganpa` is excluded
because there is an error when run it. `topologygsa` is excluded because it runs very slow.
`netgsa` is excluded because of the inconsistency between bioconductor versions.

```{r, eval = !file.exists("topo_res_list.RData"), results="none"}
all_tp_methods = setdiff(all_tp_methods, c("ganpa", "topologygsa", "netgsa"))
res_list = lapply(all_tp_methods, function(method) {
    print(method)
    nbea(method = method, se = se, gs = hsa.gs, grn = hsa.grn)
})
names(res_list) = all_tp_methods
```

```{r, echo = FALSE, eval = !file.exists("topo_res_list.RData")}
save(res_list, file = "topo_res_list.RData")
```


```{r, echo = FALSE}
load("topo_res_list.RData")
```


Get the list of significant pathways from each method. Note `gsRanking()` basically
extracts the result table.

```{r}
sig_path_list = lapply(res_list, function(x) {
    tb = gsRanking(x, signif.only = FALSE)
    tb$fdr = p.adjust(tb$PVAL, "BH")
    tb$GENE.SET[tb$fdr < 0.05]
})
sapply(sig_path_list, length)
```

Some methods generate no significant pathways under fdr < 0.05.

We compare significant pathways for all methods:

```{r}
library(eulerr)
plot(euler(sig_path_list), quantities = TRUE)
```

The results shows the results from different methods actually varies a lot.

If we also compare to GSEA method.

```{r}
res_gsea = sbea(method = "gsea", se = se, gs = hsa.gs)
res_gsea_tb = gsRanking(res_gsea, signif.only = FALSE)
res_gsea_tb$fdr = p.adjust(res_gsea_tb$PVAL, "BH")
sig_pathway_gsea = res_gsea_tb$GENE.SET[res_gsea_tb$fdr < 0.05]
length(sig_pathway_gsea)
```

Unfortunately, there is no significant pathways.

