---
title: "day4_simplifyEnrichment"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, fig.width = 8)
```

In this example, we demonstrate simplification of an ORA enrichment analysis.

```{r}
load("demo_ora.RData")
```

The following function converts gene symbols to Entrez IDs.

```{r}
convert_to_entrez_id = function(x) {
	if(is.matrix(x)) {
		map = cola:::guess_id_mapping(rownames(x))
		if(is.function(map)) {
		    new_rn = map(rownames(x))
		} else {
		    new_rn = map[rownames(x)]
		}
		l = is.na(new_rn)

		x = x[!l, , drop = FALSE]
		new_rn = new_rn[!l]

		x2 = do.call(rbind, tapply(1:nrow(x), new_rn, function(ind) {
			colMeans(x[ind, , drop = FALSE])
		}))
		return(x2)

	} else if(is.numeric(x)) {
		map = cola:::guess_id_mapping(names(x))
		x2 = s
		if(is.function(map)) {
		    names(x2) = map(names(x))
		} else {
		    names(x2) = map[names(x)]
		}
		x2 = x2[!is.na(names(x2))]
		x2 = tapply(x2, names(x2), mean)
		return(x2)
	} else {
	    map = cola:::guess_id_mapping(x)
	    if(is.function(map)) {
	        x2 = map(x)
	    } else {
	        x2 = map[x]
	    }
	    x2 = x2[!is.na(x2)]
	    x2 = unique(x2)
	    return(x2)
	}
}
diff_gene = convert_to_entrez_id(diff_gene)
```

We perform ORA analysis with **clusterProfiler**, and extract the list of significant GO terms:

```{r}
library(clusterProfiler)
library(org.Hs.eg.db)
res = enrichGO(gene = diff_gene, ont = "BP", OrgDb = org.Hs.eg.db)
go_list = res@result$ID[res@result$p.adjust < 0.05]
length(go_list)
```

Now we use **simplifyEnrichment** to simplify the reading of the massive number of GO terms.

The use of **simplifyEnrichment** is very simple. There are two steps:

1. calculate semantic similarity
2. perform clustering and visualization

```{r}
library(simplifyEnrichment)
sim_mat = GO_similarity(go_list)
simplifyGO(sim_mat)
```

```{r, echo = FALSE}
se_opt$verbose = FALSE
```

It seems there are too many clusters. There are some clusters actually showing some intermediate similarity. We can adjust the cutoff to decide whether combine or split sub-matrices.

The function `select_cutoff()` tries a list of different cutoffs and plots a diagnostic plot for deciding the optimized cutoff.

```{r}
select_cutoff(sim_mat)
```

It seems 0.92 or 0.93 is a good one, it decrease the number of clusters, although the difference score decrease a little bit (this means mean similarity within clusters decreases a little bit).

It is also interesting to see when cutoff >= 0.94, all GO terms will be put into a single cluster.


```{r}
simplifyGO(sim_mat, control = list(cutoff = 0.92))
```
------------------------------

Next we compare multiple GSEA results with **simplifyEnrichment**. We simply generate a ORA results for up-regulated genes
and a second ORA results for down-regulated genes.


```{r}
load("gene_diff_score.RData")
head(gene_diff_score)
plot(sort(gene_diff_score))
```

Setting `pvalueCutoff = 1` is very important here.


```{r}
res_up = enrichGO(gene = names(gene_diff_score)[gene_diff_score > 0.2], 
    ont = "BP", OrgDb = org.Hs.eg.db, pvalueCutoff = 1)
res_down = enrichGO(gene = names(gene_diff_score)[gene_diff_score < -0.2], 
    ont = "BP", OrgDb = org.Hs.eg.db, pvalueCutoff = 1)
```

Extract the "real" result data frames.

```{r}
tb_up = res_up@result
tb_down = res_down@result
```

Simply pass the two data frames to `simplifyGOFromMultipleLists()`. Note in the function, the column
for GO IDs and the column for FDR are automatically detected.

```{r}
simplifyGOFromMultipleLists(list(up = tb_up, down = tb_down))
```
