---
title: "day4_visualization"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, fig.height = 8)
```

We first demonstrate the visualization on ORA analysis.

```{r}
load("demo_ora.RData")
```

The following function converts gene symbols to Entrez IDs.

```{r}
convert_to_entrez_id = function(x) {
	if(is.matrix(x)) {
		map = cola:::guess_id_mapping(rownames(x))
		if(is.function(map)) {
		    new_rn = map(rownames(x))
		} else {
		    new_rn = map[rownames(x)]
		}
		l = is.na(new_rn)

		x = x[!l, , drop = FALSE]
		new_rn = new_rn[!l]

		x2 = do.call(rbind, tapply(1:nrow(x), new_rn, function(ind) {
			colMeans(x[ind, , drop = FALSE])
		}))
		return(x2)

	} else if(is.numeric(x)) {
		map = cola:::guess_id_mapping(names(x))
		x2 = s
		if(is.function(map)) {
		    names(x2) = map(names(x))
		} else {
		    names(x2) = map[names(x)]
		}
		x2 = x2[!is.na(names(x2))]
		x2 = tapply(x2, names(x2), mean)
		return(x2)
	} else {
	    map = cola:::guess_id_mapping(x)
	    if(is.function(map)) {
	        x2 = map(x)
	    } else {
	        x2 = map[x]
	    }
	    x2 = x2[!is.na(x2)]
	    x2 = unique(x2)
	    return(x2)
	}
}
diff_gene = convert_to_entrez_id(diff_gene)
```

We perform ORA analysis with **clusterProfiler**. Note the function `enrichGO()` returns a `enrichResult` object
and its slot ("`result`) contains the result table. Here we also extract the result table for future use.

Also note by default `enrichGO()` only returns significant results, so here we set `pvalueCutoff = 1` to let
`enrichGO()` to return all results (both significant and insignificant).

```{r}
library(clusterProfiler)
library(org.Hs.eg.db)
res = enrichGO(gene = diff_gene, ont = "BP", OrgDb = org.Hs.eg.db, pvalueCutoff = 1)
tb = res@result
```

How many signfiicant GO terms (FDR < 0.05)?

```{r}
sum(tb$p.adjust < 0.05)
```

First we make barplot for the significant terms:

```{r}
library(enrichplot)
barplot(res, showCategory = 20)
```


In the result table, there are fractions of geneset gene in the diff gene list and fractions of geneset genes in the background:

```{r}
head(tb)
```

We can calculate a "log2 fold change" which measures the relative value of fractions of geneset genes in the diff list and in the background.

The fraction is represented as strings "10/20", so we need to parse them into R expression:

```{r}
gene_ratio = sapply(tb$GeneRatio, function(x) eval(parse(text = x)))
bg_ratio = sapply(tb$BgRatio, function(x) eval(parse(text = x)))
log2fc = log2(gene_ratio/bg_ratio)
```

It might be interesting to see the distribution of log2 fold change.

```{r}
hist(log2fc)
```

For the ease of use, `log2fc` is appended to both `res@result` and `tb`:

```{r}
res@result$log2fc = log2fc
tb$log2fc = log2fc
```

Now we can set `log2fc` as values on x-axis in the barplot:

```{r}
barplot(res, showCategory = 20, x = "log2fc")
```

Making dot plot is almost the same as barplot. In dot plot, gene sets are by default reordered by values on x-axis.

Note x-axis starts from the minimal on x-axis.

```{r}
dotplot(res, showCategory = 20)
dotplot(res, showCategory = 20, x = "log2fc")
```

Network visualization. Note by default it only shows top 5 gene sets.

```{r}
cnetplot(res)
```


Heatmap-like plot to visualize gene-geneset relations:

```{r, fig.width=10}
heatplot(res, showCategory = 10)
```

Tree-like visualization. We first need to calculate pairwise similarity between GO terms (by default Jaccard coefficient).

```{r}
res = pairwise_termsim(res)
treeplot(res)
```

Enrichment map visualization.

```{r, fig.width=10}
emapplot(res)
```

UpSet plot.

```{r, fig.width = 16}
upsetplot(res)
```


Volcano plot:

```{r}
plot(tb$log2fc, -log10(tb$p.adjust))
abline(h = 3, lty = 2)  # FDR < 0.001
abline(v = 1, lty = 2)   # log2fc > 1
```

-------------------------------

Next we demonstrate visualization on GSEA results. First we perform GSEA also by **clusterProfiler**.
Note here we also set `pvalueCutoff = 1` because we need the full set of results for some plots.

```{r}
load("gene_diff_score.RData")
gene_diff_score = sort(gene_diff_score, decreasing = TRUE)
res_gsea = gseGO(geneList=gene_diff_score, OrgDb = org.Hs.eg.db, ont = "BP", pvalueCutoff = 1)
```

Let's check what columns are there in the result table:

```{r}
head(res_gsea@result)
```


In barplot, `NES` can be used for x-axis:

```{r}
enrichplot:::barplot.enrichResult(res_gsea, showCategory = 20, x = "NES")
dotplot(res_gsea, showCategory = 20, x = "NES")
```

Also, in Upset plot, now the distribution of gene-level statistics is added to each combination set.

```{r, fig.width = 12}
upsetplot(res_gsea)
```

Ridge plot visualizes the general distribution of gene-level statistics in each set.

```{r}
ridgeplot(res_gsea)
```

And the famous GSEA plot.

```{r}
gseaplot(res_gsea, geneSetID = 1)
```

```{r}
gseaplot2(res_gsea, geneSetID = 1)
```

We can also compare multiple gene set in one GSEA plot. In the following plot, we 
visualize the most up-regulated gene set and the most down-regulated gene set.

```{r}
ind = c(which.max(res_gsea@result$NES), which.min(res_gsea@result$NES))
gseaplot2(res_gsea, geneSetID = ind)
```


Note to make the vocalno plot. NES scores are used for x-axis:

```{r}
tb_gsea = res_gsea@result
plot(tb_gsea$NES, -log10(tb_gsea$p.adjust))
abline(h = -log10(0.05), lty = 2)  # FDR < 0.05
abline(v = c(-1, 1), lty = 2)   # log2fc > 1, log2fc < -1
```


If the GSEA is done by other packages, e.g. by **EnrichmentBrowser** package:

```{r}
library(EnrichmentBrowser)
library(CePa)
condition = read.cls("P53.cls", treatment = "MUT", control = "WT")$label
expr = read.gct("P53_collapsed_symbols.gct")

library(SummarizedExperiment)
se = SummarizedExperiment(assays = SimpleList(expr = expr))
colData(se) = DataFrame(GROUP = ifelse(condition == "WT", 1, 0))

l = condition == "WT"

library(genefilter)
tdf = rowttests(expr, factor(condition))
rowData(se) = DataFrame(FC = log2(rowMeans(expr[, l])/rowMeans(expr[, !l])),
                        ADJ.PVAL = p.adjust(tdf$p.value))

library(msigdbr)
gs = msigdbr(category = "H")
gs = split(gs$gene_symbol, gs$gs_name)

res = sbea(method = "gsea", se = se, gs = gs)
tb = gsRanking(res, signif.only = FALSE)
```

Plot the top 10 gene sets:

```{r}
top = order(tb$PVAL)[1:10]

#stat = -log10(tb$PVAL)
stat = tb$NES
names(stat) = tb$GENE.SET
barplot(stat[top], horiz = TRUE,col = ifelse(tb$PVAL[top] < 0.05, "red", "grey"))
```

Volcano plot:

```{r}
plot(tb$NES, -log10(tb$PVAL))
```
