---
title: "day4_simplifyEnrichment"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, fig.width = 8)
```

In this example, we demonstrate simplification of an GSEA enrichment analysis.

```{r}
load("gene_diff_score.RData")
gene_diff_score = sort(gene_diff_score, decreasing = TRUE)

library(clusterProfiler)
library(org.Hs.eg.db)
res = gseGO(geneList=gene_diff_score, OrgDb = org.Hs.eg.db, ont = "BP", pvalueCutoff = 1)
```


We extract the list of significant GO terms:

```{r}
go_list = res@result$ID[res@result$p.adjust < 0.1]
length(go_list)
```

Now we use **simplifyEnrichment** to simplify the reading of the massive number of GO terms.

The use of **simplifyEnrichment** is very simple. There are two steps:

1. calculate semantic similarity
2. perform clustering and visualization

```{r}
library(simplifyEnrichment)
sim_mat = GO_similarity(go_list)
simplifyGO(sim_mat)
```

------------------------------

Next we compare multiple GSEA results with **simplifyEnrichment**. We simply generate a ORA results for up-regulated genes
and a second ORA results for down-regulated genes.


```{r}
load("gene_diff_score.RData")
head(gene_diff_score)
plot(sort(gene_diff_score))
```

Setting `pvalueCutoff = 1` is very important here.


```{r}
res_up = enrichGO(gene = names(gene_diff_score)[gene_diff_score > 0.2], 
    ont = "BP", OrgDb = org.Hs.eg.db, pvalueCutoff = 1)
res_down = enrichGO(gene = names(gene_diff_score)[gene_diff_score < -0.2], 
    ont = "BP", OrgDb = org.Hs.eg.db, pvalueCutoff = 1)
```

Extract the "real" result data frames.

```{r}
tb_up = res_up@result
tb_down = res_down@result
```

Simply pass the two data frames to `simplifyGOFromMultipleLists()`. Note in the function, the column
for GO IDs and the column for FDR are automatically detected.

```{r}
simplifyGOFromMultipleLists(list(up = tb_up, down = tb_down))
```
