---
title: "Topic 2-05: Use clusterProfiler for GSEA"
author: "Zuguang Gu <z.gu@dkfz.de>"
date: '`r Sys.Date()`'
output:
  html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Topic 2-05: Use clusterProfiler for GSEA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


**Note in clusterProfiler, p-values are calculated by gene permutations!**

We use a dataset `gene_diff_score` which is a vector of a certain metric for differential expression of genes.

```{r}
gene_diff_score = readRDS(system.file("extdata", "gene_diff_score.rds", package = "GSEAtraining"))
head(gene_diff_score)
```

We need to convert genes from symbols to EntreZ IDs.

```{r}
library(org.Hs.eg.db)
map = mapIds(org.Hs.eg.db, keys = names(gene_diff_score), keytype = "SYMBOL", column = "ENTREZID")
head(map)

# names(gene_diff_score) = map[ names(gene_diff_score) ]
```

For this dataset, the mapping is 1:1. But you need to be careful when the mapping is not 1:1.

Here we use the `convert_to_entrez_id()` function from the **GSEAtraining** package, which automatically takes care of multiple mappings:


```{r}
library(GSEAtraining)
gene_diff_score = convert_to_entrez_id(gene_diff_score)
head(gene_diff_score)
```

Load the **clusterProfiler** package.

```{r, message=FALSE}
library(clusterProfiler)
```

Next we perform GSEA on different gene set sources.

Before we start, `gene_diff_score` must be pre-sorted.

```{r}
gene_diff_score = sort(gene_diff_score, decreasing = TRUE)
```

## GO enrichment

```{r, message = FALSE}
library(org.Hs.eg.db)
tb = gseGO(geneList = gene_diff_score, ont = "BP", OrgDb = org.Hs.eg.db)
head(tb)
```

## KEGG enrichment


```{r, message = FALSE}
tb = gseKEGG(geneList = gene_diff_score, organism = "hsa")
head(tb)
```

## Reactome enrichment

```{r}
library(ReactomePA)
tb = gsePathway(geneList = gene_diff_score, organism = "human")
head(tb)
```

## DO enrichment

```{r}
library(DOSE)
tb = gseDO(geneList = gene_diff_score)
head(tb)
```

## MSigDB enrichment

There is no built-in function specific for MSigDB gene sets, but there is a universal function `GSEA()` which accepts
manually-specified gene sets. The gene sets object is simply a two-column data frame:

- the first column is the gene set ID
- the second column is the gene ID

```{r}
library(msigdbr)
gene_sets = msigdbr(category = "H")
map = gene_sets[, c("gs_name", "entrez_gene")]
map$entrez_gene = as.character(map$entrez_gene)

tb = GSEA(geneList = gene_diff_score, TERM2GENE = map)
head(tb)
```


## Example for other organism

1. GO enrichment

`gseGO()` accepts an `OrgDb` object as the source of the GO gene sets.
We take pig as an example.

```{r}
library(org.Ss.eg.db)
all_genes = keys(org.Ss.eg.db, keytype = "ENTREZID")
scores = sort(rnorm(length(all_genes)), decreasing = TRUE)
names(scores) = sample(all_genes)
```


```{r, message = FALSE}
tb = gseGO(geneList = scores, ont = "BP", OrgDb = org.Ss.eg.db, pvalueCutoff = 1)
head(tb)
```


2. KEGG enrichment

The KEGG code of a specific organism can be found at https://rest.kegg.jp/list/organism

```{r, message = FALSE}
tb = gseKEGG(geneList = scores, organism = "ssc", pvalueCutoff = 1)
head(tb)
```

3. MSigDB

Use `msigdbr::msigdbr_species()` to see which organisms are supported.

```{r}
gene_sets = msigdbr(species = "pig", category = "H")
map = gene_sets[, c("gs_name", "entrez_gene")]
map$entrez_gene = as.character(map$entrez_gene)

tb = GSEA(geneList = scores, TERM2GENE = map, pvalueCutoff = 1)
head(tb)
```

## Examples for less well-studied organisms


### When an `OrgDb` object is avaiable

Taking dolphin as an example. Since `org_db` is an `OrgDb` object, we can directly use `gseGO()`:

```{r}
library(AnnotationHub)
ah = AnnotationHub()
org_db = ah[["AH112418"]]

all_genes = keys(org_db, keytype = "ENTREZID")
scores = sort(rnorm(length(all_genes)), decreasing = TRUE)
names(scores) = sample(all_genes)

tb = gseGO(geneList = scores, ont = "BP", OrgDb = org_db, pvalueCutoff = 1)
head(tb)
```


### gseKEGG() also supports many other orgainsms

```r
gseKEGG(geneList = scores, organism = ...)
```

### Manually construct the gene sets


We have introduced many ways to obtain gene sets for less well-studies organisms.
The only thing to do here is to convert gene sets to a two-column data frame where gene sets 
are in the first column and genes are in the second column. Then use `GSEA()`.

```r
GSEA(geneList = scores, TERM2GENE = ...)
```


## Look at the `tb` object

We used the same variable name `tb` for the object returned by the various enrichment functions. They are all in the same format. It looks like a table, but be careful, it is actually not:

```{r}
class(tb)
str(tb)
```

<style>
blockquote {
    padding: 10px 20px;
    margin: 0 0 20px;
    font-size: 14px;
    background: #eee;
    border-left: 6px #CCCCCC solid;
}
</style>

> In the ORA analysis by **clusterProfiler**, we mentioned using `tb@result` returns the full enrichment table. **But this is not the case for the GSEA anlaysis!**. In GSEA anlaysis 
> by **clusterProfiler**, `tb@result` still returns the significant terms. 
> Add `pvalueCutoff = 1` argument if you need the full table.


## Practice 

### {.tabset}

#### Practice 1


The input for e.g. `gseGO()` is a vector of gene-level scores, thus what metric is used for
measuring gene-level differentiation has impact on the GSEA analysis. Here you use the following
four gene-level metrics, apply them with `gseGO()` (GO BP + human) and compare the GSEA results. 

Here we use the `p53_expr.rds` dataset.

```{r}
lt = readRDS(system.file("extdata", "p53_expr.rds", package = "GSEAtraining"))
expr = lt$expr
condition = lt$condition
```

First convert gene IDs to EntreZ IDs for `expr`:

```{r}
mat = convert_to_entrez_id(expr)
```

Then calculate the column indicies of WT and MUT

```{r}
ind_MUT = which(condition == "MUT")
ind_WT = which(condition == "WT")
```

The four gene-level metrics:

1. signal-to-noise ratio
2. t-value
3. log2 fold change
4. signed log10 p-values

They can be calculated as:

```{r}
s2n = apply(mat, 1, function(x) {
    x1 = x[ind_MUT]
    x2 = x[ind_WT]

    (mean(x1) - mean(x2))/(sd(x1) + sd(x2))
})

tvalue = apply(mat, 1, function(x) {
    x1 = x[ind_MUT]
    x2 = x[ind_WT]

    t.test(x1, x2)$statistic
})

log2fc = apply(mat, 1, function(x) {
    x1 = x[ind_MUT]
    x2 = x[ind_WT]

    log2(mean(x1)/mean(x2))
})

logp = apply(mat, 1, function(x) {
    x1 = x[ind_MUT]
    x2 = x[ind_WT]

    t = t.test(x1, x2)
    sign(t$statistic)*(-log10(t$p.value))
})
```

Now use these four numeric vectors to perform GSEA analysis (on GO BP gene sets).

#### Solution

First to sort all the numeric vectors:

```{r}
s2n = sort(s2n, decreasing = TRUE)
tvalue = sort(tvalue, decreasing = TRUE)
log2fc = sort(log2fc, decreasing = TRUE)
logp = sort(logp, decreasing = TRUE)
```

```{r}
tb1 = gseGO(geneList = s2n, ont = "BP", OrgDb = org.Hs.eg.db)
tb2 = gseGO(geneList = tvalue, ont = "BP", OrgDb = org.Hs.eg.db)
tb3 = gseGO(geneList = log2fc, ont = "BP", OrgDb = org.Hs.eg.db)
tb4 = gseGO(geneList = logp, ont = "BP", OrgDb = org.Hs.eg.db)
```

Get the "real data frames" for all terms.

```{r}
tb1 = tb1@result
tb2 = tb2@result
tb3 = tb3@result
tb4 = tb4@result
```


```{r}
library(eulerr)
plot(euler(list("s2n" = tb1$ID[tb1$p.adjust < 0.05],
                "tvalue" = tb2$ID[tb2$p.adjust < 0.05],
                "log2fc" = tb3$ID[tb3$p.adjust < 0.05],
                "logp" = tb4$ID[tb4$p.adjust < 0.05])),
    quantities = TRUE)
```


```{r, fig.width = 8, fig.height = 8}
par(mfrow = c(2, 2))
plot(s2n)
plot(tvalue)
plot(log2fc)
plot(logp)
```

Conclusion: do not use log2 fold change for GSEA analysis, and using signal-to-noise or t-value (or other similar statistics) is always a good choice.

