---
title: "Topic 3-03: GOseq"
author: "Zuguang Gu <z.gu@dkfz.de>"
date: '`r Sys.Date()`'
output:
  html_document:
    toc: true
    toc_float: true
---

```{r}
library(GSEAtraining)
```

The input of **goseq** is very simple. It only needs a named binary vector with values 0 or 1,
where 1 means the gene is a DE gene. 

## Remove not-expressed genes

If the first try, we read the DE results from `de.rds`, remove genes have no p-values (mainly not expressed genes or very lowly-expressed genes), and construct the binary vector.

```{r}
tb = readRDS(system.file("extdata", "de.rds", package = "GSEAtraining"))
tb = tb[tb$symbol != "", ]  # keep pc genes
tb = tb[!is.na(tb$p_value), ]
tb$fdr = p.adjust(tb$p_value, "BH")

genes = ifelse(tb$fdr < 0.05, 1, 0)
names(genes) = tb$ensembl
head(genes)
table(genes)
```

The key assumption of **goseq** is a gene being DE is biased by its length.
The distribution of $p_\mathrm{DE}$ against gene length can be estimated by `nullp()`. (Em.. this distribution looks different from the one in the original paper)

```{r, message = FALSE}
library(goseq)
pwf = nullp(genes, "hg19", "ensGene")
```

Then use `goseq()` function to perform the test by correcting the "gene length bias":

```{r, message = FALSE}
tb1 = goseq(pwf, "hg19", "ensGene")
head(tb1)
```

## Keep all genes

This time we keep the genes with p-values as `NA` (they are basically genes not expressed).
We simply assume they are not DE.

```{r}
tb = readRDS(system.file("extdata", "de.rds", package = "GSEAtraining"))
tb = tb[tb$symbol != "", ]  # keep pc genes

tb$fdr = p.adjust(tb$p_value, "BH")
genes = ifelse(tb$fdr < 0.05, 1, 0)
genes[is.na(genes)] = 0
names(genes) = tb$ensembl
table(genes)
```

Let's check the "bias distribution" again.

```{r}
pwf = nullp(genes, "hg19", "ensGene")
```

This becomes interesting. It shows a reverse pattern as in the original paper. Note many lowly-expressed genes are short, so the increasing trend in the left part of the curve is from
the proportion of "expressed genes" getting higher? Or can we say actually there is no such bias
as mentioned in the goseq paper?

We perform the test:

```{r, message = FALSE}
tb2 = goseq(pwf, "hg19", "ensGene")
```

## Compare tb1 and tb2

We compare `tb1` and `tb2`:

```{r}
library(eulerr)
lt = list(
    test1 = tb1$category[p.adjust(tb1$over_represented_pvalue, "BH") < 0.05], 
    test2 = tb2$category[p.adjust(tb2$over_represented_pvalue, "BH") < 0.05]
)
plot(euler(lt), quantities = TRUE)
```

Does it mean the correction has no effect?

## Compare to ORA

We also compare to ORA.

```{r}
l = tb$fdr < 0.05; l[is.na(l)] = FALSE
diff_gene = tb$ensembl[l]
```

`diff_gene` are Ensembl IDs. We convert them to Entrez IDs:

```{r}
diff_gene = convert_to_entrez_id(diff_gene)
```

We use **clusterProfiler** to perform ORA analysis:


```{r, message = FALSE}
library(clusterProfiler)
library(org.Hs.eg.db)
tb3 = enrichGO(gene = diff_gene, ont = "BP", OrgDb = org.Hs.eg.db)
```

Now we compare ORA to GOseq results. Note `tb1` and `tb2` contains results for all three ontologies (BP, CC, MF). Here
we only need BP.

```{r}
tb1 = tb1[tb1$ontology == "BP", ]
tb2 = tb2[tb2$ontology == "BP", ]

library(eulerr)
lt = list(
    test1 = tb1$category[p.adjust(tb1$over_represented_pvalue, "BH") < 0.05], 
    test2 = tb2$category[p.adjust(tb2$over_represented_pvalue, "BH") < 0.05],
    ora = tb3$ID[p.adjust(tb3$pvalue, "BH") < 0.05]
)
plot(euler(lt), quantities = TRUE)
```


**Conclusion**: **goseq** was designed when the Possion distribution was used for DE analysis and
maybe it does not help nowadays when more advanced DE methods are used.
