---
title: "Topic 1-09: Gene ID conversion"
author: "Zuguang Gu <z.gu@dkfz.de>"
date: '`r Sys.Date()`'
output:
  html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Topic 1-09: Gene ID conversion}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


Gene ID type conversion is a very common task in gene set enrichment analysis. There are 
two types of packages for gene ID conversion: **biomaRt** which uses the Ensembl biomart web service
and **org.\*.db** family packages where the source information is from NCBI. Here we only introduce
the **org.\*.db** packages because they should be enough in applications.

We take **org.Hs.eg.db** (for human) as an example.

```{r}
library(org.Hs.eg.db)
```

Note: it is the same for the `OrgDb` objects for other orgainsms on **AnnotationHub**.

## Use the select() interface

We need the following three types of information:

- `keys`: Gene IDs in one ID type;
- `keytypes`: The name of the input ID type;
- `columns`: The name of the output ID type;

To get the valid name of ID types:

```{r}
keytypes(org.Hs.eg.db)
columns(org.Hs.eg.db)
```

### 1:1 mapping

For example, we want to convert the following two genes into Entrez IDs.

```{r}
genes = c("TP53", "MDM2")
```

1. use `select()`. The following function call can be read as "select 'ENTREZID' for the genes where their 'SYMBOL' are in 'gene'".

```{r}
map = select(org.Hs.eg.db, keys = genes, keytype = "SYMBOL", columns = "ENTREZID")
map
```

What if we convert them to Ensembl gene IDs:

```{r}
map = select(org.Hs.eg.db, keys = genes, keytype = "SYMBOL", columns = "ENSEMBL")
map
```

If you want to map to multiple ID types:

```{r}
select(org.Hs.eg.db, keys = genes, keytype = "SYMBOL", columns = c("ENTREZID", "ENSEMBL"))
```

2. use `mapIds()` which is very similar to `select()`.

Note the argument is named `column` instead of `columns`, so you can only map to one gene ID type.


```{r}
mapIds(org.Hs.eg.db, keys = genes, keytype = "SYMBOL", column = "ENTREZID")
mapIds(org.Hs.eg.db, keys = genes, keytype = "SYMBOL", column = "ENSEMBL")
```


### 1:many mapping

Now there might be some problems if the mapping is not 1:1.

```{r}
genes = c("TP53", "MMD2")
map = select(org.Hs.eg.db, keys = genes, keytype = "SYMBOL", columns = "ENTREZID")
map
```

Usually it is hard to pick one unique gene for such 1:mapping case, but we can add an additional
column "GENETYPE" when querying:

```{r}
map = select(org.Hs.eg.db, keys = genes, keytype = "SYMBOL", columns = c("ENTREZID", "GENETYPE"))
map
```

For "MMD2", adding the "GENETYPE" column works because the second hit of it is annotated to "unknown".
We can simply remove it.

```{r}
map[map$GENETYPE == "protein-coding", ]
```

And it is always a good idea to only inlucde protein-coding genes in gene set enrichment analysis.


But for `mapIds()`, it is not as flexible as `select()`. The `multiVals` argument controls how to select
the gene when the mapping is not 1-vs-1.

```{r}
mapIds(org.Hs.eg.db, keys = genes, keytype = "SYMBOL", column = "ENTREZID", multiVals = "first")
mapIds(org.Hs.eg.db, keys = genes, keytype = "SYMBOL", column = "ENTREZID", multiVals = "asNA")
mapIds(org.Hs.eg.db, keys = genes, keytype = "SYMBOL", column = "ENTREZID", multiVals = "list")
mapIds(org.Hs.eg.db, keys = genes, keytype = "SYMBOL", column = "ENTREZID", multiVals = "filter")

gene_type = as.list(org.Hs.egGENETYPE)
# this is very slow
mapIds(org.Hs.eg.db, keys = genes, keytype = "SYMBOL", column = "ENTREZID", 
    multiVals = function(x) {
        x2 = x[sapply(as.list(gene_type[x]), function(y) y == "protein-coding")]
        if(length(x2)) {
            x2[1]
        } else {
            NA
        }
})
```


## Use the pre-generated objects

In **org.\*.db**, there are also pre-generated objects that already contains mapping between EntreZ IDs to a specific gene ID type.

```{r}
ls(envir = asNamespace("org.Hs.eg.db"))
```

The following six objects can be used to convert between major gene ID types:

- `org.Hs.egENSEMBL`: Entrez -> Ensembl
- `org.Hs.egENSEMBL2EG`: Ensembl -> Entrez
- `org.Hs.egREFSEQ`: Entrez -> RefSeq
- `org.Hs.egREFSEQ2EG`: RefSeq -> Entrez
- `org.Hs.egSYMBOL`: Entrez -> Symbol
- `org.Hs.egSYMBOL2EG`: Symbol -> Entrez

```{r}
org.Hs.egSYMBOL
```

If you have a single gene, you can use `[[`:

```{r}
org.Hs.egSYMBOL2EG[["TP53"]]
```


If you have multiple genes, use `[` + `as.list()`:

```{r}
lt = as.list(org.Hs.egSYMBOL2EG)
lt[genes]

# or
as.list(org.Hs.egSYMBOL2EG[genes])
```

You can also use `toTable()`, but it is less used for converting gene IDs:

```{r}
tb = toTable(org.Hs.egSYMBOL2EG)
```

As you can see, the drawback is you cannot use additional information to filter genes, such as gene types.


## Conclusion

If the mappig is 1:1, all the methods metioned is OK. When the mapping is 1:many, use `select()` + filtering by the "GENETYPE" column is safer.


In the **GSEAtraining** package, there is one such function. It also automatically
detect the input gene ID type.

```{r, eval = FALSE}
library(GSEAtraining)
convert_to_entrez_id(genes)
```

## Practice

### {.tabset}

#### Practice 1

In the object `diff_gene` from the following code, there is a vector of gene symbols (human), try to convert them to EntreZ IDs (do not use `convert_to_entrez_id()`).


```{r}
lt = readRDS(system.file("extdata", "ora.rds", package = "GSEAtraining"))
diff_gene = lt$diff_gene
head(diff_gene)
```

#### Solution


```{r, eval = FALSE}
map = select(org.Hs.eg.db, keys = diff_gene, keytype = "SYMBOL", columns = "ENTREZID")
unique(map$ENTREZID)

# or
map = mapIds(org.Hs.eg.db, keys = diff_gene, keytype = "SYMBOL", column = "ENTREZID")
```



### {.tabset}

#### Practice 2

Now we know gene ID mapping between different ID types is not always 1:1. Count
how many (the percent) gene symbols can not be uniquely mapped to EntreZ IDs, and how
many (the percent) Ensembl IDs can not be uniquely mapped to EntreZ IDs. And recalculate
these two numbers only taking into account the protein-coding genes.

#### Solution


```{r}
all_symbols = keys(org.Hs.eg.db, keytype = "SYMBOL")

# use select
map = select(org.Hs.eg.db, keys = all_symbols, keytype = "SYMBOL", columns = "ENTREZID")

tb = table(map$SYMBOL)
sum(tb > 1)/length(tb)

# use mapIds
map = mapIds(org.Hs.eg.db, keys = all_symbols, keytype = "SYMBOL", column = "ENTREZID", 
    multiVals = "list")
n = sapply(map, length)
sum(n > 1)/length(n)

# use org.Hs.egSYMBOL2EG
lt = as.list(org.Hs.egSYMBOL2EG)
n = sapply(lt, length)
sum(n > 1)/length(n)
```

Only protein-coding genes:

```{r}
# use select
map = select(org.Hs.eg.db, keys = all_symbols, keytype = "SYMBOL", 
    columns = c("ENTREZID", "GENETYPE"))
map = map[map$GENETYPE == "protein-coding", ]

tb = table(map$SYMBOL)
sum(tb > 1)/length(tb)
```


**Ensembl -> Entrez**


```{r}
all_ids = keys(org.Hs.eg.db, keytype = "ENSEMBL")

# use select
map = select(org.Hs.eg.db, keys = all_ids, keytype = "ENSEMBL", columns = "ENTREZID")

tb = table(map$ENSEMBL)
sum(tb > 1)/length(tb)

tail(sort(tb))
```

Only protein-coding genes:

```{r}
# use select
map = select(org.Hs.eg.db, keys = all_ids, keytype = "ENSEMBL", columns = c("ENTREZID", "GENETYPE"))
map = map[map$GENETYPE == "protein-coding", ]

tb = table(map$ENSEMBL)
sum(tb > 1)/length(tb)
```

Last, we add a SYMBOL column in the map table and check what are these Ensembl genes which map to so many Entrez genes: 

```{r}
map = select(org.Hs.eg.db, keys = all_ids, keytype = "ENSEMBL", 
    columns = c("ENTREZID", "SYMBOL", "GENETYPE"))
map[map$ENSEMBL %in% names(tail(sort(tb))), ]
```