---
title: "Topic 1-03: Get pathways from MSigDB"
author: "Zuguang Gu <z.gu@dkfz.de>"
date: '`r Sys.Date()`'
output:
  html_document:
    toc: true
    toc_float: true
---

```{r}
library(GSEAtraining)
```

## The .gmt data format for gene sets

MSigDB defines a simple `.gmt` format for storing gene sets. It has the following format:

```
gene_set_1  gene_set_description    gene1   gene2   gene3
gene_set_2  gene_set_description    gene4   gene5
...
```

The `.gmt` format is used more and more for sharing gene set data, e.g. https://maayanlab.cloud/Enrichr/#libraries.


Let's try to read a `.gmt` file. We use the `.gmt` file for the Hallmark gene set.

```{r}
download.file("https://data.broadinstitute.org/gsea-msigdb/msigdb/release/2023.2.Hs/h.all.v2023.2.Hs.symbols.gmt",
  destfile = "h.all.v2023.2.Hs.symbols.gmt")
```

Since it is not a table, we have to read the file by lines.

```{r}
ln = readLines("h.all.v2023.2.Hs.symbols.gmt")
ln = strsplit(ln, "\t")
gs = lapply(ln, function(x) x[-(1:2)])
names(gs) = sapply(ln, function(x) x[1])
gs[1:2]
```

```{r, echo = FALSE}
invisible(file.remove("h.all.v2023.2.Hs.symbols.gmt"))
```


## The msigdbr package

There is already an R package **msigdbr** which has integerated all MSigDB gene sets.

```{r, message = FALSE}
library(msigdbr)
```

The major advantage is, MSigDB only supports human and mouse, **msigdbr** supports
more organisms by mapping to the orthologe genes.

```{r}
msigdbr_species()
```


All categories of gene sets:

```{r}
as.data.frame(msigdbr_collections())
```

Values in the `gs_cat` column and the `gs_subcat` column can be used to retrieve gene sets
in a specific category.
E.g., we want to extract genesets in C2 category and CP:KEGG sub-category:

```{r}
gene_sets = msigdbr(category = "C2", subcategory = "CP:KEGG")
gene_sets
```

If you don't like the `tibble` format, you can change it to a data frame.

```{r}
gene_sets = as.data.frame(gene_sets)
head(gene_sets)
```

<style>
blockquote {
    padding: 10px 20px;
    margin: 0 0 20px;
    font-size: 14px;
    background: #eee;
    border-left: 6px #CCCCCC solid;
}
</style>


> **Danger zone:** The `entrez_gene` column is saved in integer. This may cause unseeable problems.
> Convert it explicitely to character: `gene_set$entrez_gene = as.character(gene_set$entrez_gene)`.

Another advantage of **misgdbr** is that it provides all three major gene ID types.
For example, if you need Ensembl gene IDs, you can obtain the gene sets by:

```{r}
gene_sets[, c("gs_name", "ensembl_gene")] |> head()
```

By default it retrieves data for human. The `species` argument can be used for other supported species.

```{r}
gene_sets = msigdbr(species = "dog", category = "C2", subcategory = "CP:BIOCARTA")
gene_sets = as.data.frame(gene_sets)
head(gene_sets)
```


## A safer check

The table contains mappings between gene sets and all the three gene ID types. Since Symbol-EntreZ-Ensembl mapping is not 1:1, it is possible rows are duplicated if only taking
two columns from the complete table:

```{r}
gene_sets = msigdbr(category = "C5", subcategory = "GO:BP")

tb = gene_sets[, c("gs_name", "gene_symbol")]
dim(tb)
dim(unique(tb))
```


So, additionally applying `unique()` is always safe.

## Functions in the GSEAtraining package

We also have several functions in **GSEAtraining** to retrieve gene sets from MSigDB.
It directly downloads data from MSigDB web server.

```{r}
list_msigdb_versions()
list_msigdb_collections("2023.2.Hs")
lt = get_msigdb(version = "2023.2.Hs", collection = "h.all")
lt[1:2]
```


## Practice 

### {.tabset}

#### Practice 1

The gene set resource on https://maayanlab.cloud/Enrichr/#libraries is very useful. You may
want to use it some say in the future. Take one gene set collection (e.g. the COVID-19 related gene sets), download the corresponding gene set file (in `.gmt` format) and try to read into R as a list.

#### Solution


```{r}
download.file("https://maayanlab.cloud/Enrichr/geneSetLibrary?mode=text&libraryName=COVID-19_Related_Gene_Sets", destfile = "covid-19.gmt")
ln = readLines("covid-19.gmt")
ln = strsplit(ln, "\t")
gs = lapply(ln, function(x) x[-(1:2)])
names(gs) = sapply(ln, function(x) x[1])
gs[1:2]
```


```{r, echo = FALSE}
invisible(file.remove("covid-19.gmt"))
```
