---
title: "Topic 1-07: Get GO/KEGG gene sets for other organisms"
author: "Zuguang Gu <z.gu@dkfz.de>"
date: '`r Sys.Date()`'
output:
  html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Topic 1-07: Get GO/KEGG gene sets for other organisms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Besides those well-studied organisms, there are also resources for less well-studies organisms, mainly for KEGG and GO gene sets.


## KEGG pathways

KEGG pathway supports a huge number of organisms, just try different organism code:

```{r, eval = FALSE}
read.table(url("https://rest.kegg.jp/list/pathway/hsa"), sep = "\t")
read.table(url("https://rest.kegg.jp/list/pathway/ptr"), sep = "\t")
read.table(url("https://rest.kegg.jp/list/pathway/pps"), sep = "\t")
read.table(url("https://rest.kegg.jp/list/pathway/ggo"), sep = "\t")
read.table(url("https://rest.kegg.jp/list/pathway/pon"), sep = "\t")
...
```

## BioMartGOGeneSets

The **BioMartGOGeneSets** package has compiled GO gene sets for > 700 organisms using [the biomart web service](https://www.ensembl.org/info/data/biomart/index.html) (with the **biomaRt** package.

To obtain the gene sets, use the function `getBioMartGOGeneSets()`. Also you need to provide
the "dataset". The name of the dataset can be obtained by `supportedOrganisms()` or from https://bioconductor.org/packages/devel/data/annotation/vignettes/BioMartGOGeneSets/inst/doc/supported_organisms.html. Let's get GO gene sets
for mouse. 

```{r}
library(BioMartGOGeneSets)
lt = getBioMartGOGeneSets("mmusculus_gene_ensembl")
length(lt)
lt[1]
```

The variable `lt` is a list of vectors where each vector corresponds to a GO gene set with Ensembl
IDs as gene identifiers.

You can try the following command and see what will be printed:

```{r, eval = FALSE}
lt = getBioMartGOGeneSets("mouse")
```

In `getBioMartGOGeneSets()`, argument `as_table` can be set to `TRUE`, then the function returns
a data frame.

```{r}
tb = getBioMartGOGeneSets("mmusculus_gene_ensembl", as_table = TRUE)
head(tb)
```

Argument `ontology` controls which category of GO gene sets. Possible values should be `"BP"`, `"CC"`
and `"MF"`.

```{r, eval = FALSE}
getBioMartGOGeneSets("mmusculus_gene_ensembl", ontology = "BP") # the default one
getBioMartGOGeneSets("mmusculus_gene_ensembl", ontology = "CC")
getBioMartGOGeneSets("mmusculus_gene_ensembl", ontology = "MF")
```

Last, argument `gene_id_type` can be set to `"entrez_gene"` or `"gene_symbol"`, then genes in the gene sets
are in Entrez IDs or gene symbols. Note this depends on specific organisms, that not every organism supports 
Entrez IDs or gene symbols.

```{r}
lt = getBioMartGOGeneSets("mmusculus_gene_ensembl", gene_id_type = "entrez_gene")
lt[1]

lt = getBioMartGOGeneSets("mmusculus_gene_ensembl", gene_id_type = "gene_symbol")
lt[1]
```

## OrgDb objects

Biocondutor core team maintaines **org.\*.db** packages for 18 organisms

```{r, echo = FALSE}
tb = read.table(textConnection(
"org.Hs.eg.db    Human   org.Ss.eg.db    Pig
org.Mm.eg.db    Mouse   org.Gg.eg.db    Chicken
org.Rn.eg.db    Rat org.Mmu.eg.db   Rhesus_monkey
org.Dm.eg.db    Fruit_fly   org.Cf.eg.db    Canine
org.At.tair.db  Arabidopsis org.EcK12.eg.db E_coli_strain_K12
org.Sc.sgd.db   Yeast   org.Xl.eg.db    African_clawed_frog
org.Dr.eg.db    Zebrafish   org.Ag.eg.db    Malaria_mosquito
org.Ce.eg.db    Nematode    org.Pt.eg.db    Chimpanzee
org.Bt.eg.db    Bovine  org.EcSakai.eg.db   E_coli_strain_Sakai
"))
```

```{r, echo = FALSE}
tb[, 1] = paste0("`", tb[, 1], "`")
tb[, 3] = paste0("`", tb[, 3], "`")
tb[, 2] = gsub("_", " ", tb[, 2])
tb[, 4] = gsub("_", " ", tb[, 4])
knitr::kable(tb, col.names = c("Package", "Organism", "Package", "Organism"))
```

These **org.\*.db** packages can be used in the same way as what we have seen with the **org.Hs.eg.db** package.

On **AnnotationHub**, there are `OrgDb` objects for a huge number of other organisms (~2000) which can be used
for getting GO genes and mappings between gene IDs.

```{r}
library(AnnotationHub)
ah = AnnotationHub()
```

To search for an organism, using its latin name is more suggested. Also add the "OrgDb" (data class) keyword:

```{r}
query(ah, c("cat", "OrgDb"))
```

```{r}
query(ah, c("Felis catus", "OrgDb"))
```


Besides using `query()` to search for the AnnotationHub dataset, you can
also use the **BiocHubsShiny** package to interactively searching for datasets:

Now we download the data.

```{r}
org_db = ah[["AH111810"]]  # using `[[` downloads the dataset
```


`org_db` is an `OrgDb` object but contains less information than the **org.\*.db** packages:

```{r}
org_db
columns(org_db)
```

You can obtain the GO gene sets manually by `select()`, taking the `"ENTREZID"` and `"GOALL"` columns:

```{r}
all_genes = keys(org_db, keytype = "ENTREZID")
tb = select(org_db, keys = all_genes, keytype = "ENTREZID", columns = c("GOALL", "ONTOLOGYALL"))
head(tb)
```

You may need to clean the mapping table:

```{r}
tb = tb[!is.na(tb$GOALL), ]
tb = unique(tb)
```



With an `OrgDb` object, you can also use the helper function `get_GO_gene_sets_from_orgdb()` in **GSEAtraining** package:

```{r}
library(GSEAtraining)
lt = get_GO_gene_sets_from_orgdb(org_db, "BP")
lt[1]
```



## Practice 

### {.tabset}

#### Practice 1

Try to obtain GO (BP) gene sets for dolphin (latin name: Tursiops truncatus) from **BioMartGOGeneSets**
and **AnnotationHub**. Compare the difference of the GO gene sets from these two sources, e.g.
compare the number of annotated genes from the two sources.

#### Solution

Get GO gene sets from biomart. First search the "dataset name" with `supportedOrganisms()`.

```{r}
gs1 = getBioMartGOGeneSets("ttruncatus_gene_ensembl", "BP", gene_id_type = "entrez_id")
```


Get GO gene sets from NCBI/AnnotationHub.

```{r}
query(ah, c("Tursiops truncatus", "OrgDb"))
orgdb = ah[["AH112418"]]
gs2 = get_GO_gene_sets_from_orgdb(orgdb, "BP")
```


```{r}
cn = intersect(names(gs1), names(gs2))
n1 = sapply(gs1, length)
n2 = sapply(gs2, length)

plot(n1[cn], n2[cn], xlab = "getBioMartGOGeneSets", ylab = "AnnotationHub")
```

Actually, the GO gene sets from `getBioMartGOGeneSets()`/Ensembl is from an older genome build for dolphin. So it is more suggested to use the data from AnnotationHub/NCBI.
