---
title: "Topic 2-03: Use clusterProfiler for ORA"
author: "Zuguang Gu <z.gu@dkfz.de>"
date: '`r Sys.Date()`'
output:
  html_document:
    toc: true
    toc_float: true
---

```{r}
library(GSEAtraining)
```

Let's start with a vector of DE genes:

```{r}
lt = readRDS(system.file("extdata", "ora.rds", package = "GSEAtraining"))
diff_gene = lt$diff_gene
bg_gene = lt$bg_gene
head(diff_gene)
```

Load the **clusterProfiler** package.

```{r, message=FALSE}
library(clusterProfiler)
```

**clusterProfiler** supports several major gene ID types in the input, but it is suggested
to use Entrez IDs as the input because it is the "central gene ID type" in many databases/datasets.

We convert `diff_gene` to Entrez IDs. Note some genes may be lost 
due to the conversion.

```{r}
diff_gene = convert_to_entrez_id(diff_gene)
head(diff_gene)
length(diff_gene)
```

Next we perform ORA on different gene set sources.

## GO enrichment

```{r, message = FALSE}
library(org.Hs.eg.db)
tb = enrichGO(gene = diff_gene, ont = "BP", OrgDb = org.Hs.eg.db)
head(tb)
```

## KEGG enrichment


```{r, message = FALSE}
tb = enrichKEGG(gene = diff_gene, organism = "hsa")
head(tb)
```

## Reactome enrichment

```{r}
library(ReactomePA)
tb = enrichPathway(gene = diff_gene, organism = "human")
head(tb)
```

## DO enrichment

```{r}
library(DOSE)
tb = enrichDO(gene = diff_gene, ont = "DO")
head(tb)
```

## MSigDB enrichment

There is no built-in function specific for MSigDB gene sets, but there is a universal function `enricher()` which accepts
manually-specified gene sets. The gene sets object is simply a two-column data frame:

- the first column is the gene set ID
- the second column is the gene ID

```{r}
library(msigdbr)
gene_sets = msigdbr(category = "H")
map = gene_sets[, c("gs_name", "entrez_gene")]
head(map)

tb = enricher(gene = diff_gene, TERM2GENE = map)
head(tb)
```

## Setting background

Note, in **clusterProfiler**, the background is intersect(all genes in the gene sets, user's background genes).

```{r}
bg_gene = convert_to_entrez_id(bg_gene)
bg_gene = sample(bg_gene, 10000)
tb = enrichGO(gene = diff_gene, ont = "BP", OrgDb = org.Hs.eg.db, universe = bg_gene)
head(tb)
```

We can compare ORA with different backgrounds:

```{r}
tb1 = enrichGO(gene = diff_gene, ont = "BP", OrgDb = org.Hs.eg.db)
tb2 = enrichGO(gene = diff_gene, ont = "BP", OrgDb = org.Hs.eg.db, universe = bg_gene)

tb1 = tb1@result
tb2 = tb2@result
cn = intersect(tb1$ID, tb2$ID)

plot(tb1[cn, "pvalue"], tb2[cn, "pvalue"], log = "xy",
	xlim = c(1e-25, 1), ylim = c(1e-25, 1),
	xlab = "default background", ylab = "self-defined background")
```

Or overlap between the significant GO terms:

```{r}
library(eulerr)
plot(euler(list("default_background" = tb1$ID[tb1$p.adjust < 0.05],
	            "bg_gene" = tb2$ID[tb2$p.adjust < 0.05])),
    quantities = TRUE)
```

## Example for other organism

1. GO enrichment

`enrichGO()` accepts an `OrgDb` object as the source of the GO gene sets.
We take pig as an example.

```{r}
library(org.Ss.eg.db)
diff_gene = random_genes(org.Ss.eg.db, 1000, "ENTREZID")
```


```{r, message = FALSE}
tb = enrichGO(gene = diff_gene, ont = "BP", OrgDb = org.Ss.eg.db, 
	pvalueCutoff = 1, qvalueCutoff = 1)
head(tb)
```


2. KEGG enrichment

The KEGG code of a specific organism can be found at https://rest.kegg.jp/list/organism

```{r, message = FALSE}
tb = enrichKEGG(gene = diff_gene, organism = "ssc", pvalueCutoff = 1, qvalueCutoff = 1)
head(tb)
```

3. MSigDB

Use `msigdbr::msigdbr_species()` to see which organisms are supported.

```{r}
gene_sets = msigdbr(species = "pig", category = "H")
map = gene_sets[, c("gs_name", "entrez_gene")]
map$entrez_gene = as.character(map$entrez_gene)

tb = enricher(gene = diff_gene, TERM2GENE = map, pvalueCutoff = 1, qvalueCutoff = 1)
head(tb)
```

## Examples for less well-studied organisms

Taking dolphin as an example. 

```{r}
library(AnnotationHub)
ah = AnnotationHub()
org_db = ah[["AH112418"]]
gs = get_GO_gene_sets_from_orgdb(org_db, "BP", as_table = TRUE)

diff_gene = random_genes(org_db, 1000, "ENTREZID")
tb = enricher(diff_gene, TERM2GENE = gs[, 2:1], pvalueCutoff = 1, qvalueCutoff = 1)
head(tb)
```

And we can add GO full names to the table:

```{r}
library(GO.db)
tb = tb@result
tb$Name = Term(GOTERM[tb$ID])
head(tb)
```

or 

```{r, eval = FALSE}
enricher(diff_gene, TERM2GENE = gs[, 2:1], TERM2NAME = toTable(GOTERM)[, 2:3],
	pvalueCutoff = 1, qvalueCutoff = 1)
```


Since `org_db` is an `OrgDb` object, we can directly use `enrichGO()`:

```{r}
tb = enrichGO(diff_gene, ont = "BP", OrgDb = org_db, 
	pvalueCutoff = 1, qvalueCutoff = 1)
```

## Look at the `tb` object

We used the same variable name `tb` for the object returned by the various enrichment functions. They are all in the same format. It looks like a table, but be careful, it is actually not:

```{r}
class(tb)
str(tb)
```

To convert it into a "real" table, DO NOT use `as.data.frame()` which only returns the significant terms, use `tb@result`.


## Practice 

### {.tabset}

#### Practice 1

Use the DE gene list in `webgestalt_example_gene_list.rds` to perform ORA on
GO/KEGG/MSigDB gene sets.

```{r}
genes = readRDS(system.file("extdata", "webgestalt_example_gene_list.rds", 
	package = "GSEAtraining"))
```

