---
title: "Topic 2-06: Compare ORA and GSEA"
author: "Zuguang Gu <z.gu@dkfz.de>"
date: '`r Sys.Date()`'
output:
  html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Topic 2-06: Compare ORA and GSEA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


We apply ORA and GSEA on the same p53 datasets and compare the results.

```{r}
lt = readRDS(system.file("extdata", "p53_expr.rds", package = "GSEAtraining"))
expr = lt$expr
condition = lt$condition
```

We convert gene IDs to Entrez IDs in `expr`:

```{r}
library(GSEAtraining)
expr = convert_to_entrez_id(expr)
```

**ORA**

We need to perform t-test or similar DE test to extract DE genes:

We apply t-test to every gene.

```{r}
p = apply(expr, 1, function(x) {
    x1 = x[condition == "WT"]
    x2 = x[condition == "MUT"]
    
    t.test(x1, x2)$p.value
})
```


Faster row-wise t-test: 

```{r}
library(genefilter)
tdf = rowttests(expr, factor(condition))  # the second must be a "factor"
tdf$fdr = p.adjust(tdf$p.value, "BH")
sum(tdf$fdr < 0.05)  # number of diff genes
```

It seems there are not too many diff genes. We look at the distribution of the t-statistics:

```{r}
plot(sort(tdf$statistic))
```

Instead of setting a cutoff for FDR, we set a cutoff for t-statistics, just to get
enough number of diff genes.

```{r}
sum(abs(tdf$statistic) > 2)
```

First we perform ORA, using GO BP gene sets:

```{r, message = FALSE}
library(clusterProfiler)
library(org.Hs.eg.db)
l_sig = abs(tdf$statistic) > 2
tb_ora = enrichGO(gene  = rownames(expr)[l_sig],
                  OrgDb = org.Hs.eg.db,
                  ont   = "BP",
                  pAdjustMethod = "BH")
tb_ora = as.data.frame(tb_ora)  # only significant ones
```


Then we perform GSEA, using t-statistics as gene scores. 

```{r}
s = tdf$statistic
names(s) = rownames(tdf)   # s must have names (gene IDs)
s = sort(s, decreasing = TRUE)  # s must be pre-sorted
tb_gsea = gseGO(geneList = s, 
                OrgDb = org.Hs.eg.db,
                ont   = "BP",
                pAdjustMethod = "BH")
tb_gsea = as.data.frame(tb_gsea)  # only significant ones
```

We compare the two significant GO lists:

```{r}
library(eulerr)
plot(euler(list(ORA = tb_ora$ID, GSEA = tb_gsea$ID)), quantities = TRUE)
```
