---
title: "Topic 6-01: Visualization"
author: "Zuguang Gu <z.gu@dkfz.de>"
date: '`r Sys.Date()`'
output:
  html_document:
    toc: true
    toc_float: true
---

```{r}
library(GSEAtraining)
```

## ORA result

### Single ORA result

```{r}
lt = readRDS(system.file("extdata", "ora.rds", package = "GSEAtraining"))
diff_gene = lt$diff_gene
diff_gene = convert_to_entrez_id(diff_gene)
```

```{r}
library(clusterProfiler)
library(org.Hs.eg.db)
res = enrichGO(gene = diff_gene, ont = "BP", OrgDb = org.Hs.eg.db, pvalueCutoff = 1)
res = add_more_columns(res)
```

Let's start with the enrichment table from the ORA analysis. 

```{r}
tb = res@result
head(tb)
```

The most used type of plot is to use simple statistical graphics for a small set of pre-selected gene sets.

```{r}
library(ggplot2)
ggplot(tb[1:10, ], aes(x = n_hits, y = Description)) + 
    geom_col()
```

Number of DE genes in gene set may not be a good statistic, more commonly used statistics
are log2 fold enrichment or -log10 p.adjust.

ggplot(tb[1:10, ], aes(x = n_hits, y = factor(Description, levels = Description))) + geom_col()

```{r}
ggplot(tb[1:10, ], aes(x = log2_fold_enrichment, y = Description)) + 
    geom_col()
```

It is also common to add the p-values/adjusted p-values to the bars.

```{r}
ggplot(tb[1:10, ], aes(x = log2_fold_enrichment, y = Description)) + 
    geom_col() +
    geom_text(aes(x = log2_fold_enrichment, label = sprintf('%1.e', p.adjust)))
```

By default, **ggplot2** reorders labels alphabetically, 

```{r}
ggplot(tb[1:10, ], aes(x = log2_fold_enrichment, y = factor(Description, levels = Description))) + 
    geom_col()
```

Points are also very frequently used.

```{r}
ggplot(tb[1:10, ], aes(x = log2_fold_enrichment, y = Description)) + 
    geom_point() +
    xlim(0, max(tb$log2_fold_enrichment[1:10]))
```

Using dot plot, we can map a second statistic to the size of dots.

```{r}
ggplot(tb[1:10, ], aes(x = log2_fold_enrichment, y = Description, size = n_hits)) + 
    geom_point() +
    xlim(0, max(tb$log2_fold_enrichment[1:10]))
```

Even more, we can map a third statistic to the dot colors.

```{r}
ggplot(tb[1:10, ], aes(x = gs_size, y = Description, size = n_hits, col = n_hits/gs_size)) + 
    geom_point()

```


A volcano plot which is -log10(p.adjust) vs log2 fold enrichment:

```{r}
ggplot(tb, aes(x = log2_fold_enrichment, y = -log10(p.adjust))) +
    geom_point(col = ifelse(tb$log2_fold_enrichment > 1 & tb$p.adjust < 0.001, "black", "grey")) +
    geom_hline(yintercept = 3, lty = 2) + geom_vline(xintercept = 1, lty = 2)
```

With the volcano plot, we can easily see the preference of enrichment to the sizes of gene sets.

```{r}
ggplot(tb, aes(x = log2_fold_enrichment, y = -log10(p.adjust), color = gs_size, size = n_hits)) +
    geom_point() + scale_colour_distiller(palette = "Spectral")

```

### Visualizations from clusterProfiler

**clusterProfiler** provides many visualization functions (now these functions are all moved
into a new **enrichplot** package).

```{r, fig.height = 10}
library(enrichplot)
cnetplot(res)
```

```{r}
heatplot(res, showCategory = 10)
```

```{r, fig.height = 9}
res = pairwise_termsim(res)
treeplot(res)
```

```{r, fig.width = 10, fig.height = 10}
emapplot(res)
```

```{r, fig.height = 6, fig.width = 10}
upsetplot(res)
```

### Multiple ORA results

To visualize multiple ORA enrichment tables in one plot, we need to first prepare a data frame
which combines results for a pre-selected gene sets.

```{r}
lt = readRDS(system.file("extdata", "lt_enrichment_tables.rds"))

set.seed(666)
terms = sample(lt[[1]]$ID, 10)

tb = NULL
for(nm in names(lt)) {
    x = lt[[nm]]
    x = x[x$ID %in% terms, colnames(x) != "geneID"]
    x$sample = nm
    tb = rbind(tb, x)
}
```

```{r}
ggplot(tb, aes(x = sample, y = Description, size = -log10(p.adjust))) +
    geom_point(color = ifelse(tb$p.adjust < 0.05, "black", "grey"))
```

```{r}
ggplot(tb, aes(x = sample, y = wrap_text(Description), size = -log10(p.adjust))) +
    geom_point(color = ifelse(tb$p.adjust < 0.05, "black", "grey"))
```

## GSEA result

### Single GSEA result

```{r}
gene_diff_score = readRDS(system.file("extdata", "gene_diff_score.rds", package = "GSEAtraining"))
gene_diff_score = convert_to_entrez_id(gene_diff_score)
gene_diff_score = sort(gene_diff_score, decreasing = TRUE)
res_gsea = gseGO(geneList = gene_diff_score, ont = "BP", OrgDb = org.Hs.eg.db, pvalueCutoff = 1)
tb = res_gsea@result
head(tb)
```

Similarly, we can visualize NES scores of gene sets.

```{r}
ggplot(tb[1:10, ], aes(x = NES, y = Description)) + 
    geom_col()
ggplot(tb[1:10, ], aes(x = NES, y = Description)) + 
    geom_col(fill = ifelse(tb$NES[1:10] > 0, "red", "darkgreen"))
```

The volcano plot:

```{r}
ggplot(tb, aes(x = NES, y = -log10(p.adjust))) +
    geom_point(col = ifelse(abs(tb$NES) > 1 & tb$p.adjust < 0.05, "black", "grey")) +
    geom_hline(yintercept = -log10(0.05), lty = 2) + geom_vline(xintercept = c(1, -1), lty = 2)
```

The ridge plot (joy plot/mountain plot) visualizes the distribution of gene-level scores in each gene set.

```{r, fig.height = 10}
ridgeplot(res_gsea)
```

And the classic GSEA plot:

```{r}
gseaplot(res_gsea, geneSetID = 1)
gseaplot2(res_gsea, geneSetID = 1)
```

```{r}
ind = c(which.max(res_gsea@result$NES), which.min(res_gsea@result$NES))
gseaplot2(res_gsea, geneSetID = ind)
```