---
title: "Topic 2-04: Implement GSEA from scratch"
author: "Zuguang Gu <z.gu@dkfz.de>"
date: '`r Sys.Date()`'
output:
  html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Topic 2-04: Implement GSEA from scratch}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


In this document, we will implement the two versions of GSEA in R, using the p53 dataset
also from the GSEA paper.

```{r}
lt = readRDS(system.file("extdata", "p53_expr.rds", package = "GSEAtraining"))
expr = lt$expr
condition = lt$condition

ln = strsplit(readLines(system.file("extdata", "c2.symbols.gmt", package = "GSEAtraining")), "\t")
gs = lapply(ln, function(x) x[-(1:2)])
names(gs) = sapply(ln, function(x) x[1])

geneset = gs[["p53hypoxiaPathway"]]
```

This gene set is very small:

```{r}
length(geneset)
```



Note here gene IDs in the expression matrix and in the gene set are all gene
symbols, thus no more adjustment needs to be done here.


The gene-level difference score is set as signal-to-noise ratios, which is:

- mean in group 1
- mean in group 2
- sd in group 1
- sd in group 2

$$ \frac{\mu_1 - \mu_2}{\sigma_1 + \sigma_2} $$

We calculate the gene-level difference score in `s`:

```{r}
s = apply(expr, 1, function(x) {
	x1 = x[condition == "WT"]
	x2 = x[condition == "MUT"]
	(mean(x1) - mean(x2))/(sd(x1) + sd(x2))
})
```

A faster way to calculate `s` is to use the **matrixStats** packages.

```r
library(matrixStats)
m1 = expr[, condition == "WT"]  # only samples in group 1
m2 = expr[, condition == "MUT"]  # only samples in group 2
s = (rowMeans(m1) - rowMeans(m2))/(rowSds(m1) + rowSds(m2)) # gene-level difference scores
```

Sort the gene scores from the highest to the lowest:

```{r}
s = sort(s, decreasing = TRUE)
```

## GSEA version 1

### Calculate enrichment score

Next we first implement the original GSEA method, which was proposed in Mootha et al., 2003.


We calculate the cummulative probability of a gene being in a gene set $G_k$. At position $p$ in the
sorted gene score vector:

$$ f_{1p} = \frac{1}{n_k} \sum_{i=1}^p I(g_i \in G_k) $$


```{r}
## original GSEA
l_set = names(s) %in% geneset
f1 = cumsum(l_set)/sum(l_set)

## or
binary_set = l_set + 0
f1 = cumsum(binary_set)/sum(binary_set)
```

Here in the calculation of `f1`, `sum(l_set)` is $n_k$, `cumsum(l_set)` is $\sum_{i=1}^p I(g_i \in G_k)$.

Similarly, we calculate the cummulative distribution of a gene not in the gene set.

```{r}
l_other = ! names(s) %in% geneset
f2 = cumsum(l_other)/sum(l_other)
```

`f1` and `f2` can be used to plot the CDFs of two distributions.

```{r}
n = length(s)
plot(1:n, f1, type = "l", col = "red", xlab = "sorted genes")
lines(1:n, f2, col = "blue")
```

The reason why the blue locates almost on the diagonal is the gene set is very small.

Next the difference of cumulative probability (`f1 - f2`) at each position of the sorted gene list.
Let's call it "the GSEA plot".

```{r}
plot(f1 - f2, type = "l", xlab = "sorted genes")
abline(h = 0, lty = 2, col = "grey")
points(which(l_set), rep(0, sum(l_set)), pch = "|", col = "red")
```

So, a large value of `f1 - f2` means the genes are more like to be in the gene set for those locating at `1..p`.

The enrichment score (ES) defined as `max(f1 - f2)` is:

```{r}
es = max(f1 - f2)
es
```

And the position in the "GSEA plot":

```{r}
plot(f1 - f2, type = "l", xlab = "sorted genes")
abline(h = 0, lty = 2, col = "grey")
points(which(l_set), rep(0, sum(l_set)), pch = "|", col = "red")
abline(v = which.max(f1 - f2), lty = 3, col = "blue")
```

The statistic `es` actually is the Kolmogorov-Smirnov statistics, thus, we can
directly apply the KS test where the two vectors for the KS test are the positions
on the sorted gene list.

```{r}
ks.test(which(l_set), which(l_other))
```

However, we can see the p-value is not significant, this is because KS test is not
a powerful test. Next we construct the null distribution by sample permutation.

### Sample permutation

In the next code chunk, the calculation of ES score is wrapped into a function,
also we use `rowMeans()` and `rowSds()` to speed up the calculation of gene-level
scores.

```{r}
library(matrixStats)
# expr: the complete expression matrix
# condition: the condition labels of samples
# cmp: a vector of two, cmp[1] - cmp[2] > 0 means up-regulation
# geneset: A vector of genes
calculate_es = function(expr, condition, cmp, geneset) {

	m1 = expr[, condition == cmp[1]]  # only samples in group 1
	m2 = expr[, condition == cmp[2]]  # only samples in group 2

	s = (rowMeans(m1) - rowMeans(m2))/(rowSds(m1) + rowSds(m2)) # a gene-level difference socre (S2N ratio) 

	s = sort(s, decreasing = TRUE)  # ranked gene list

	l_set = names(s) %in% geneset
	f1 = cumsum(l_set)/sum(l_set)   # CDF for genes in the set

	l_other = !l_set
	f2 = cumsum(l_other)/sum(l_other)  # CDF for genes not in the set

	max(f1 - f2)
}
```

The ES score calculated by `calculate_es()`:

```{r}
es = calculate_es(expr, condition, cmp = c("WT", "MUT"), geneset = geneset)
es
```

We randomly permute sample labels or we randomly permute `condition`. These two ways
are identical. The aim is to break the association between `condition` and `expr`.

```r
condition_random = sample(condition)
expr = expr[, sample(ncol(expr))]
```

We do sample permutation
1000 times. The random ES scores are saved in `es_rand`.

```{r}
set.seed(123)
es_rand = numeric(1000)
for(i in 1:1000) {
	es_rand[i] = calculate_es(expr, sample(condition), 
	    cmp = c("WT", "MUT"), geneset = geneset)
}
```

p-value is calculated as the proportion of values in `es_rand` being equal to or larger than `es`.

```{r}
sum(es_rand >= es)/1000
```


The null distribution of ES:

```{r}
hist(es_rand)
abline(v = es, col = "red")
```

## GSEA version 2

Next we implement the improved GSEA (Subramanian et al., PNAS, 2005) where gene-level scores are taken as the weight.

We directly modify `calculate_es()` to `calculate_es_v2()` where there is only two lines new, which we highlight in the code chunk:

```{r}
calculate_es_v2 = function(expr, condition, cmp, geneset, plot = FALSE, power = 1) {

    m1 = expr[, condition == cmp[1]]
    m2 = expr[, condition == cmp[2]]

    s = (rowMeans(m1) - rowMeans(m2))/(rowSds(m1) + rowSds(m2))

    s = sort(s, decreasing = TRUE)

    l_set = names(s) %in% geneset
    
    # f1 = cumsum(l_set)/sum(l_set)  # <<-- the original line
    
    s_set = abs(s)^power   # <<-- here
    s_set[!l_set] = 0
    f1 = cumsum(s_set)/sum(s_set)  ## <<- here

    l_other = !l_set
    f2 = cumsum(l_other)/sum(l_other)

    if(plot) {
        plot(f1 - f2, type = "l", xlab = "sorted genes")
        abline(h = 0, lty = 2, col = "grey")
        points(which(l_set), rep(0, sum(l_set)), pch = "|", col = "red")
        abline(v = which.max(f1 - f2), lty = 3, col = "blue")
    }

    max(f1 - f2)
}
```

Now we calculate the new ES score and make the GSEA plot:

```{r}
es = calculate_es_v2(expr, condition, cmp = c("WT", "MUT"), plot = TRUE, 
    geneset = geneset)
```

We can also check when `power = 0` and `power = 2`:

```{r, results = "none", fig.width = 10}
par(mfrow = c(1, 2))
calculate_es_v2(expr, condition, cmp = c("WT", "MUT"), plot = TRUE, power = 0, 
    geneset = geneset)  # same as the original GSEA
title("power = 0")
calculate_es_v2(expr, condition, cmp = c("WT", "MUT"), plot = TRUE, power = 2, 
    geneset = geneset)
title("power = 2")
```

Similarly, we randomly permute samples to obtain the null distribution of ES:

```{r}
es_rand = numeric(1000)
for(i in 1:1000) {
	es_rand[i] = calculate_es_v2(expr, sample(condition), 
	    cmp = c("WT", "MUT"), geneset = geneset)
}
```

The new p-value:

```{r}
sum(es_rand >= es)/1000
```

The minimal p-value from 1000 permutations is 1/1000, so zeor here means p-value < 0.001.

And the null distribution of ES:

```{r}
hist(es_rand, xlim = c(0, 1))
abline(v = es, col = "red")
```

We can see the improved GSEA is more powerful than the original GSEA, because the 
original GSEA equally weights genes and the improved GSEA weights genes based on their differential expression,
which increases the effect of diff genes. Let's plot the weight of genes:

```{r}
plot(abs(s))
```


## GSEA version 2, gene permutation

Null distribution can also be constructed by gene permutation. It is very
easy to implement:

```{r}
# s: a vector of pre-calcualted gene-level scores
# s should be sorted
calculate_es_v2_gene_perm = function(s, geneset, perm = FALSE, plot = FALSE, power = 1) {
	
	if(perm) {
	    # s is still sorted, but the gene labels are randomly shuffled
	    # to break the associations between gene scores and gene labels.
		names(s) = sample(names(s))  ## <<- here
	}

	l_set = names(s) %in% geneset
	s_set = abs(s)^power
	s_set[!l_set] = 0
	f1 = cumsum(s_set)/sum(s_set)

	l_other = !l_set
	f2 = cumsum(l_other)/sum(l_other)

	if(plot) {
        plot(f1 - f2, type = "l", xlab = "sorted genes")
        abline(h = 0, lty = 2, col = "grey")
        points(which(l_set), rep(0, sum(l_set)), pch = "|", col = "red")
        abline(v = which.max(f1 - f2), lty = 3, col = "blue")
    }

	max(f1 - f2)
}
```

Good thing of gene permutation is the gene-level scores only need to be calculated
once and can be repeatedly used.


```{r}
# pre-calculate gene-level scores
m1 = expr[, condition == "WT"]
m2 = expr[, condition == "MUT"]

s = (rowMeans(m1) - rowMeans(m2))/(rowSds(m1) + rowSds(m2))
s = sort(s, decreasing = TRUE)  # must be pre-sorted
```

The GSEA plot under gene permutation:

```{r}
es = calculate_es_v2_gene_perm(s, geneset, plot = TRUE)
```

We calculate the null distribution of ES from gene permutation:


```{r}
es_rand = numeric(1000)
for(i in 1:1000) {
	es_rand[i] = calculate_es_v2_gene_perm(s, geneset, perm = TRUE)
}

sum(es_rand >= es)/1000
```

The real p-value is also < 0.001.

The null distribution of ES from gene permutation:

```{r}
hist(es_rand, xlim = c(0, 1))
abline(v = es, col = "red")
```

## Other statistics

Taking `es` and `es_rand` calculated previously, NES is calculated as

```{r}
es/mean(es_rand)
```


NES does not take into account of the dispersion of `es_rand`, sometimes z-scores is used instead:

```{r}
(es - mean(es_rand))/sd(es_rand)
```


## Practice 

### {.tabset}

#### Practice 1


We have demonstrated two different ways (sample permutation and gene permutation) for calculating p-values. Apply the p53 dataset on the 50 hallmark gene sets, and compare the two enrichment results (e.g. to check which method tends to generate more signficant p-values).

The code we saw so far only applies to a single gene set, you may need to put it into a `for` loop to
go over each of the 50 hallmark gene sets.


#### Solution

The Hallmark gene sets from MSigDB

```{r}
library(GSEAtraining)
gs_hallmark = get_msigdb(version = "2023.2.Hs", collection = "h.all", gene_id_type = "symbol")
n_gs = length(gs_hallmark)
```

Sample permutation:

```{r}
p1 = numeric(n_gs)
es1 = numeric(n_gs)
nes1 = numeric(n_gs)
z1 = numeric(n_gs)

for(i in 1:n_gs) {
	print(i)
	geneset = gs_hallmark[[i]]

	es1[i] = calculate_es_v2(expr, condition, 
		    cmp = c("WT", "MUT"), geneset = geneset)

	es_rand = numeric(1000)
	for(k in 1:1000) {
		es_rand[k] = calculate_es_v2(expr, sample(condition), 
		    cmp = c("WT", "MUT"), geneset = geneset)
	}
	p1[i] = sum(es_rand >= es1[i])/1000
	nes1[i] = es1[i]/mean(es_rand)
	z1[i] = (es1[i] - mean(es_rand))/sd(es_rand)
}

df1 = data.frame(
	geneset = names(gs_hallmark),
	es = es1,
	nes = nes1,
	z = z1,
	p_value = p1
)
```


Gene permutation:

```{r}
m1 = expr[, condition == "WT"]
m2 = expr[, condition == "MUT"]

s = (rowMeans(m1) - rowMeans(m2))/(rowSds(m1) + rowSds(m2))
s = sort(s, decreasing = TRUE)  # must be pre-sorted

p2 = numeric(n_gs)
es2 = numeric(n_gs)
nes2 = numeric(n_gs)
z2 = numeric(n_gs)

for(i in 1:n_gs) {
	print(i)
	geneset = gs_hallmark[[i]]

	es2[i] = calculate_es_v2_gene_perm(s, geneset = geneset)
	
	es_rand = numeric(1000)
	for(k in 1:1000) {
		es_rand[k] = calculate_es_v2_gene_perm(s, geneset = geneset, perm = TRUE)
	}
	p2[i] = sum(es_rand >= es2[i])/1000
	nes2[i] = es2[i]/mean(es_rand)
	z2[i] = (es2[i] - mean(es_rand))/sd(es_rand)
}

df2 = data.frame(
	geneset = names(gs_hallmark),
	es = es2,
	nes = nes2,
	z = z2,
	p_value = p2
)
```

Compare ES, NES, z-scores, and p-values:

```{r, fig.width = 10, fig.height = 10}
par(mfrow = c(2, 2))
plot(df1$es, df2$es, main = "ES",
	xlab = "sample permutation", ylab = "gene permutation")
plot(df1$nes, df2$nes, main = "NES", xlim = c(0, 2.5), ylim = c(0, 2.5),
	xlab = "sample permutation", ylab = "gene permutation")
plot(df1$z, df2$z, main = "Z-score",
	xlab = "sample permutation", ylab = "gene permutation")
plot(df1$p_value, df2$p_value, main = "P-value",
	xlab = "sample permutation", ylab = "gene permutation")
```
