---
title: "Topic 4-01: CePa"
author: "Zuguang Gu <z.gu@dkfz.de>"
date: '`r Sys.Date()`'
output:
  html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Topic 4-01: CePa}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


## CePa

The **CePa** package has not been updated for ~10 years. I am planning to revisit it
some time this year. 


It will try fix different centralities:

- equal.weight
- in.degree
- out.degree
- node betweenness
- in.reach
- out.reach

```{r}
library(CePa)
data(PID.db)
data(gene.list)
```

`dif` and `bk` are two gene vectors, `pc` is in a special format.

```{r, eval = FALSE}
# around 5min
res = cepa.ora.all(dif = gene.list$dif, bk = gene.list$bk, pc = PID.db$NCI)
```

```{r, echo = FALSE}
res = readRDS(system.file("extdata", "cepa_res.rds", package = "GSEAtraining"))
```

Heatmap of adjusted p-values:

```{r}
plot(res, adj.method = "BH", only.sig = TRUE)
```

We can see "e2f_pathway" is signinicant with in-degree but not significant with out-degree.
Let's directly plot the pathway network:

Plot network for a specific pathway under a specific centrality

```{r}
plot(res, "e2f_pathway", cen = "in.degree")
plot(res, "e2f_pathway", cen = "out.degree")
```

