---
title: "Topic 1-05: Get pathways from BioCarta"
author: "Zuguang Gu <z.gu@dkfz.de>"
date: '`r Sys.Date()`'
output:
  html_document:
    toc: true
    toc_float: true
---

```{r}
library(GSEAtraining)
```

[BioCarta](https://web.archive.org/web/20170122225118/https://cgap.nci.nih.gov/Pathways/BioCarta_Pathways) is a valuable source of biological pathways which not only provides well manually curated pathways, but also remarkable and intuitive pathway images.

## The BioCartaImage package

The **BioCartaImage** is mainly for customizing BioCarta pathway images, but of source it contains gene sets for BioCarta pathways.

```{r}
library(BioCartaImage)
head(PATHWAY2ENTREZ)
```

The full name of pathways:

```{r}
pathway_names = sapply(BIOCARTA_PATHWAYS, function(x) x$name)
head(pathway_names)
```


### Highlight genes on the pathway image

```{r}
library(grid)
grid.newpage()
grid.biocarta("h_RELAPathway", color = c("1387" = "yellow"))
```


Add more complicated graphics:

```{r}
grob = biocartaGrob("h_RELAPathway")
grob2 = mark_gene(grob, "1387", function(x, y) {
    pos = pos_by_polygon(x, y)
    pushViewport(viewport(x = pos[1] - 10, y = pos[2], 
        width = unit(4, "cm"), height = unit(4, "cm"), 
        default.units = "native", just = "right"))
    grid.rect(gp = gpar(fill = "red"))
    grid.text("add whatever\nyou want here")
    popViewport()
}, capture = TRUE)
grid.draw(grob2)
```