---
title: "Topic 1-06: Get gene sets from UniProt Keywords"
author: "Zuguang Gu <z.gu@dkfz.de>"
date: '`r Sys.Date()`'
output:
  html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Topic 1-06: Get gene sets from UniProt Keywords}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


UniProt database provides a list of controlled vocabulary represented as
keywords for genes or proteins (https://www.uniprot.org/keywords/). This is
useful for summarizing gene functions in a compact way.

## The UniProtKeywords package

First load the package:

```{r}
library(UniProtKeywords)
```

The release and source information of the data:

```{r}
UniProtKeywords
```


The **UniProtKeywords** package has compiled genesets of keywords for some
organism, which can get by the function `load_keyword_genesets()`. The argument
is the taxon ID of a organism. The full set of supported organisms can be found
in the document of `load_keyword_genesets()` (or in the object `UniProtKeywords:::ORGANISM`).

```{r}
gl = load_keyword_genesets(9606)
gl[3:4]  # because gl[1:2] has a very long output, here we print gl[3:4]
```

You can also use the name of the organism:

```{r, eval = FALSE}
load_keyword_genesets("human")
load_keyword_genesets("Homo sapiens")
```

Argument `as_table` can be set to `TRUE`, then `load_keyword_genesets()` returns a two-column data frame.

```{r}
tb = load_keyword_genesets(9606, as_table = TRUE)
head(tb)
```

## Statistics

We can simply check some statistics.

1. Sizes of keyword genesets:


```{r, fig.width = 7, fig.height = 5}
plot(table(sapply(gl, length)), log = "x", 
    xlab = "Size of keyword genesets",
    ylab = "Number of keywords"
)
```

2. Numbers of words in keywords:

```{r, fig.width = 7, fig.height = 5}
plot(table(sapply(gregexpr(" |-|/", names(gl)), length)), 
    xlab = "Number of words in keywords",
    ylab = "Number of keywords"
)
```

3. Numbers of characters in keywords:

```{r, fig.width = 7, fig.height = 5}
plot(table(nchar(names(gl))), 
    xlab = "Number of characters in keywords",
    ylab = "Number of keywords"
)
```