---
title: "Topic 4-02: GSVA"
author: "Zuguang Gu <z.gu@dkfz.de>"
date: '`r Sys.Date()`'
output:
  html_document:
    toc: true
    toc_float: true
---

```{r}
library(GSEAtraining)
```

## GSVA

We need an expression matrix as well as a collection of gene sets.

```{r}
lt = readRDS(system.file("extdata", "p53_expr.rds", package = "GSEAtraining"))
expr = lt$expr
condition = lt$condition

ln = strsplit(readLines(system.file("extdata", "c2.symbols.gmt", package = "GSEAtraining")), "\t")
gs = lapply(ln, function(x) x[-(1:2)])
names(gs) = sapply(ln, function(x) x[1])
```

Running **GSVA** analysis is simple. Just note the gene IDs in `expr` should match the gene IDs in `gs`.

The returned value `gs_mat` is a set-sample matrix that contains "single sample-based gene set variation scores".

```{r}
library(GSVA)
gs_mat = gsva(expr, gs, verbose = FALSE)
dim(gs_mat)
```


`gs_mat` can be used for downstream analysis. E.g. make heatmaps:


```{r, message = FALSE}
library(ComplexHeatmap)
Heatmap(gs_mat, top_annotation = HeatmapAnnotation(cond = condition),
    column_split = condition)
```


Apply t-test on each row of `gs_mat` to test whether the gene set-level
profile has difference between the two conditions.

```{r, message=FALSE}
library(genefilter)
tdf = rowttests(gs_mat, factor(condition))
tdf$fdr = p.adjust(tdf$p.value, "BH")
```

How many gene sets are significant?

```{r}
tdf[tdf$fdr < 0.05, ]
```

## Compare to normal GSEA

As a comparison, we also perform a normal GSEA analysis. We use the t-value as the gene-level score:

```{r}
s = rowttests(expr, factor(condition))[, "statistic"]
names(s) = rownames(expr)
```

Here we also test to the c2 gene set collection, so we need to convert `gs` to the format **clusterProfiler** accepts:

```{r, message = FALSE}
map = data.frame(
    gene_set = rep(names(gs), times = sapply(gs, length)),
    gene = unlist(gs)
)
    
library(clusterProfiler)
tb = GSEA(geneList = sort(s, decreasing = TRUE), TERM2GENE = map, pvalueCutoff = 1)
head(tb)
```


Let's check the expression of genes in the p53Pathway:

```{r}
g = intersect(gs[["p53Pathway"]], rownames(expr))
mm = expr[g, ]
mm = t(scale(t(mm)))  # z-score transformation

Heatmap(mm, name = "z-score", top_annotation = HeatmapAnnotation(cond = condition),
    column_title = "p53Pathway",
    right_annotation = rowAnnotation(bar = anno_barplot(s[g], axis_param = list(direction = "reverse"), width = unit(2, "cm"))),
    column_split = condition) %v%
Heatmap(gs_mat["hivnefPathway", , drop = FALSE], name = "set variation",
    col = circlize::colorRamp2(c(-1, 0, 1), c("green", "yellow", "red")))
```

Next we make pairwise scatterplot of p-values and statistics for both analysis.

```{r, fig.width = 10, fig.height = 5}
cn = intersect(rownames(tdf), tb$ID)
par(mfrow = c(1, 2))
plot(tdf[cn, "p.value"], tb[cn, "pvalue"],
    xlab = "GSVA", ylab = "normal GSEA", main = "p-values")
plot(tdf[cn, "statistic"], tb[cn, "NES"],
    xlab = "GSVA / t-values", ylab = "normal GSEA / NES")
```

This makes a problem here that one method generates a significant gene set
which can be completely insignificant under another method (left plot). The
direction of the general differentiation of a gene set can be reversed in
different methods.

```{r}
g = intersect(gs[["hivnefPathway"]], rownames(expr))
mm = expr[g, ]
mm = t(scale(t(mm)))  # z-score transformation

Heatmap(mm, name = "z-score", top_annotation = HeatmapAnnotation(cond = condition),
    right_annotation = rowAnnotation(bar = anno_barplot(s[g], axis_param = list(direction = "reverse"), width = unit(2, "cm"))),
    column_split = condition) %v%
Heatmap(gs_mat["hivnefPathway", , drop = FALSE], name = "set variation",
    col = circlize::colorRamp2(c(-1, 0, 1), c("green", "yellow", "red")))
```


**Conclusion**: It does not seem ssGSEA is better than GSEA. Use with caution.

