---
title: "Topic 2-02: Implement ORA from stratch"
author: "Zuguang Gu <z.gu@dkfz.de>"
date: '`r Sys.Date()`'
output:
  html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Topic 2-02: Implement ORA from stratch}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---



Implementing ORA is very simple. As long as genes and gene sets are already in the correct ID types and formats, the implementation of ORA only needs several lines.


Three inputs for ORA:

- a vector of (DE) genes: `genes`
- a list of vectors where each vector contains genes in a gene set: `gene_sets`
- a vector of background genes: `universe`

We calculate the following numbers:

Number of total genes: 

```r
n_universe = length(universe)
```

Number of DE genes:

```r
n_genes = length(genes)
```

Sizes of gene sets:

```r
m = sapply(gene_sets, length)
```


Number of DE genes in gene sets:

```r
x = sapply(gene_sets, function(x) length(intersect(x, genes)))
```

We rename `n_genes` to `k`:

```r
k = n_genes
```


<style>
.table {
width:400px;
}
</style>

We put these variables into the 2x2 contigency table:

|      | in the set | not in the set| total |
|------|------------:|---------------:|-------:|
|DE    |  `x`    |   -   |  `k` |
|not DE|  -    |      -   |  - |
|total |  `m`    |    -     |  `n_universe` |


Then we can calculate p-values from the hypergeometric distribution as:

```r
phyper(x - 1, m, n_universe - m, k, lower.tail = FALSE)
```

which is the same as:

```r
phyper(x - 1, k, n_universe - k, m, lower.tail = FALSE)
```

Let's put all these code into a function:


```r
ora = function(genes, gene_sets, universe) {
    n_universe = length(universe)
    n_genes = length(genes)
    
    x = sapply(gene_sets, function(x) length(intersect(x, genes)))
    m = sapply(gene_sets, length)
    n = n_universe - m
    k = n_genes
    
    p = phyper(x - 1, m, n, k, lower.tail = FALSE)

    data.frame(
        gene_set = names(gene_sets),
        hits = x,
        n_genes = k,
        n_gs = m,
        n_total = n_universe,
        p_value = p
    )
}
```

We can further improve the function by:

1. set the default of `universe` to the total genes in `gene_sets`
2. intersect `genes` and `gene_sets` to `universe`. Note this step also removes duplicated genes
3. add adjusted log2 fold enrichment and p-values
4. order the result table by the significance

```{r}
ora = function(genes, gene_sets, universe = NULL) {

    if(is.null(universe)) {
        universe = unique(unlist(gene_sets))
    } else {
        universe = unique(universe)
    }
    # restrict in universe
    genes = intersect(genes, universe)
    gene_sets = lapply(gene_sets, function(x) intersect(x, universe))

    n_universe = length(universe)
    n_genes = length(genes)
    
    x = sapply(gene_sets, function(x) length(intersect(x, genes)))
    m = sapply(gene_sets, length)
    n = n_universe - m
    k = n_genes
    
    p = phyper(x - 1, m, n, k, lower.tail = FALSE)

    df = data.frame(
        gene_set = names(gene_sets),
        hits = x,
        n_genes = k,
        n_gs = m,
        n_total = n_universe,
        log2fe = log2(x*n_universe/k/m),
        p_value = p
    )

    df$p_adjust = p.adjust(df$p_value, "BH")
    rownames(df) = df$gene_set
    df[order(df$p_adjust, df$p_value), ,drop = FALSE]
}
```

Let's try our `ora()` with 1000 random genes on the MSigDB hallmark gene sets.
`get_msigdb()` is from the **GSEAtraining** package.

```{r}
library(org.Hs.eg.db)
library(GSEAtraining)
gs = get_msigdb(version = "2023.2.Hs", collection = "h.all")
genes = random_genes(org.Hs.eg.db, 1000, "ENTREZID")

df = ora(genes, gs)
head(df)
```

The only restriction of `ora()` is the ID type for `genes` should be
the same as in `gene_sets`.


## Practice 

### {.tabset}

#### Practice 1

The file `de.rds` contains results from a differential expression (DE) analysis.

```{r}
de = readRDS(system.file("extdata", "de.rds", package = "GSEAtraining"))
de = de[, c("symbol", "p_value")]
de = de[!is.na(de$p_value), ]
head(de)
```

Take the `symbol` and `p_value` columns, set cutoff of `p_value` to 0.05, 0.01
and 0.001 respectively to filter the significant DE genes (of course, in
applications, you need to correct p-values), apply `ora()` to the three DE
gene lists using the GO BP gene sets (think how to obtain the GO gene sets from **org.Hs.eg.db**) and compare the ORA results (the whatever comparison
you can think of).

If you EntreZ ID in the gene sets, you need to convert gene symbols for the DE genes to EntreZ IDs.
But you can also choose to use gene symbols in the gene sets.

#### Solution

The three DE gene lists:

```{r}
de_genes_1 = de$symbol[de$p_value < 0.05]
de_genes_2 = de$symbol[de$p_value < 0.01]
de_genes_3 = de$symbol[de$p_value < 0.001]

length(de_genes_1)
length(de_genes_2)
length(de_genes_3)
```

Get the GO BP gene sets. Note genes in `de_genes_*` are symbols, thus genes in gene sets
should also be symbols.

```{r}
gs = get_GO_gene_sets_from_orgdb(org.Hs.eg.db, "BP", gene_id_type = "SYMBOL")
gs[1]
```


We perform ORA for the three DE gene lists.

```{r}
tb1 = ora(de_genes_1, gs)
tb2 = ora(de_genes_2, gs)
tb3 = ora(de_genes_3, gs)
```

Compare number of significant GO terms:

```{r}
sum(tb1$p_adjust < 0.05)
sum(tb2$p_adjust < 0.05)
sum(tb3$p_adjust < 0.05)
```

Actually we can see gene set enrichment analysis is a very robust method
which is not sensitive to DE cutoffs (at least for this dataset).

Make a Venn (Euler) diagram:


```{r}
library(eulerr)
plot(euler(list("DE_cutoff_0.05"  = tb1$gene_set[tb1$p_adjust < 0.05],
                "DE_cutoff_0.01"  = tb2$gene_set[tb2$p_adjust < 0.05],
                "DE_cutoff_0.001" = tb3$gene_set[tb3$p_adjust < 0.05])), 
     quantities = TRUE)
```

Compare log2 fold enrichment from the three ORA results. First take the common GO terms
in the three ORA tables.

```{r}
cn = intersect(intersect(rownames(tb1), rownames(tb2)), rownames(tb3))
```

Then simply make pairwise scatter plots:

```{r, fig.width = 12, fig.height = 4}
par(mfrow = c(1, 3))
plot(tb1[cn, "log2fe"], tb2[cn, "log2fe"], pch = 16, cex = 0.5, col = "#00000080",
    xlim = c(-5, 5), ylim = c(-5, 5), main = "compare log2 fold enrichment from ORA",
    xlab = "DE_cutoff_0.05", ylab = "DE_cutoff_0.01")
abline(a = 0, b = 1, lty = 2, col = "red")

plot(tb1[cn, "log2fe"], tb3[cn, "log2fe"], pch = 16, cex = 0.5, col = "#00000080",
    xlim = c(-5, 5), ylim = c(-5, 5), main = "compare log2 fold enrichment from ORA",
    xlab = "DE_cutoff_0.05", ylab = "DE_cutoff_0.001")
abline(a = 0, b = 1, lty = 2, col = "red")

plot(tb2[cn, "log2fe"], tb3[cn, "log2fe"], pch = 16, cex = 0.5, col = "#00000080",
    xlim = c(-5, 5), ylim = c(-5, 5), main = "compare log2 fold enrichment from ORA",
    xlab = "DE_cutoff_0.01", ylab = "DE_cutoff_0.001")
abline(a = 0, b = 1, lty = 2, col = "red")
```
