---
title: "Topic 2-03: Use clusterProfiler for ORA"
author: "Zuguang Gu <z.gu@dkfz.de>"
date: '`r Sys.Date()`'
output:
  html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Topic 2-03: Use clusterProfiler for ORA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


Let's start with a vector of DE genes:

```{r}
lt = readRDS(system.file("extdata", "ora.rds", package = "GSEAtraining"))
diff_gene = lt$diff_gene
bg_gene = lt$bg_gene
head(diff_gene)
```

Load the **clusterProfiler** package.

```{r, message=FALSE}
library(clusterProfiler)
```

**clusterProfiler** supports several major gene ID types in the input, but it is suggested
to use Entrez IDs as the input because it is the "central gene ID type" in many databases/datasets.

We convert `diff_gene` to Entrez IDs. Note some genes may be lost 
due to the conversion.

```{r}
library(GSEAtraining)
diff_gene = convert_to_entrez_id(diff_gene)
head(diff_gene)
length(diff_gene)
```

Next we perform ORA on different gene set sources.

## GO enrichment

```{r, message = FALSE}
library(org.Hs.eg.db)
tb = enrichGO(gene = diff_gene, ont = "BP", OrgDb = org.Hs.eg.db)
head(tb)
```

## KEGG enrichment


```{r, message = FALSE}
tb = enrichKEGG(gene = diff_gene, organism = "hsa")
head(tb)
```

## Reactome enrichment

```{r}
library(ReactomePA)
tb = enrichPathway(gene = diff_gene, organism = "human")
head(tb)
```

## DO enrichment

```{r}
library(DOSE)
tb = enrichDO(gene = diff_gene)
head(tb)
```

## MSigDB enrichment

There is no built-in function specific for MSigDB gene sets, but there is a universal function `enricher()` which accepts
manually-specified gene sets. The gene sets object is simply a two-column data frame:

- the first column is the gene set ID
- the second column is the gene ID

```{r}
library(msigdbr)
gene_sets = msigdbr(category = "H")
map = gene_sets[, c("gs_name", "entrez_gene")]
head(map)

tb = enricher(gene = diff_gene, TERM2GENE = map)
head(tb)
```

## Setting background

Note, in **clusterProfiler**, the background is intersect(all genes in the gene sets, user's background genes).

```{r}
bg_gene = convert_to_entrez_id(bg_gene)
bg_gene = sample(bg_gene, 10000)
tb = enrichGO(gene = diff_gene, ont = "BP", OrgDb = org.Hs.eg.db, universe = bg_gene)
head(tb)
```

We can compare ORA with different backgrounds:

```{r}
tb1 = enrichGO(gene = diff_gene, ont = "BP", OrgDb = org.Hs.eg.db)
tb2 = enrichGO(gene = diff_gene, ont = "BP", OrgDb = org.Hs.eg.db, universe = bg_gene)

tb1 = tb1@result
tb2 = tb2@result
cn = intersect(tb1$ID, tb2$ID)

plot(tb1[cn, "pvalue"], tb2[cn, "pvalue"], log = "xy",
	xlim = c(1e-25, 1), ylim = c(1e-25, 1),
	xlab = "default background", ylab = "self-defined background")
```

Or overlap between the significant GO terms:

```{r}
library(eulerr)
plot(euler(list("default_background" = tb1$ID[tb1$p.adjust < 0.05],
	            "bg_gene" = tb2$ID[tb2$p.adjust < 0.05])),
    quantities = TRUE)
```

## Example for other organism

1. GO enrichment

`enrichGO()` accepts an `OrgDb` object as the source of the GO gene sets.
We take pig as an example. `random_genes()` is from **GSEAtraining** package.

```{r}
library(org.Ss.eg.db)
diff_gene = random_genes(org.Ss.eg.db, 1000, "ENTREZID")
```

Here we set `pvalueCutoff = 1, qvalueCutoff = 1` because genes are random genes and
it is expected there won't be too much significant terms left.


```{r, message = FALSE}
tb = enrichGO(gene = diff_gene, ont = "BP", OrgDb = org.Ss.eg.db, 
	pvalueCutoff = 1, qvalueCutoff = 1)
head(tb)
```


2. KEGG enrichment

The KEGG code of a specific organism can be found at https://rest.kegg.jp/list/organism

```{r, message = FALSE}
tb = enrichKEGG(gene = diff_gene, organism = "ssc", pvalueCutoff = 1, qvalueCutoff = 1)
head(tb)
```

3. MSigDB

Use `msigdbr::msigdbr_species()` to see which organisms are supported.

```{r}
gene_sets = msigdbr(species = "pig", category = "H")
map = gene_sets[, c("gs_name", "entrez_gene")]
map$entrez_gene = as.character(map$entrez_gene)

tb = enricher(gene = diff_gene, TERM2GENE = map, pvalueCutoff = 1, qvalueCutoff = 1)
head(tb)
```

## Examples for less well-studied organisms

### When an `OrgDb` object is avaiable

You can use `enrichGO()` as long as you have an `OrgDb` object:

Taking dolphin as an example. 

```{r}
library(AnnotationHub)
ah = AnnotationHub()
org_db = ah[["AH112418"]]

diff_gene = random_genes(org_db, 1000, "ENTREZID")
tb = enrichGO(diff_gene, ont = "BP", OrgDb = org_db, 
	pvalueCutoff = 1, qvalueCutoff = 1)
head(tb)
```

### enrichKEGG() also supports many other orgainsms

```r
enrichKEGG(gene = diff_gene, organism = ...)
```

### Manually construct the gene sets


We have introduced many ways to obtain gene sets for less well-studies organisms.
The only thing to do here is to convert gene sets to a two-column data frame where gene sets 
are in the first column and genes are in the second column. Then use `enricher()`.

```r
enricher(diff_gene, TERM2GENE = ...)
```


## Look at the `tb` object

We used the same variable name `tb` for the object returned by the various enrichment functions. They are all in the same format. It looks like a table, but be careful, it is actually not:

```{r}
class(tb)
str(tb)
```

To convert it into a "real" table, DO NOT use `as.data.frame()` which only returns the significant terms, use `tb@result`.

Also some columns in `tb@result` are not in the proper format, e.g. `"GeneRatio"` and `"BgRatio"` where numbers should be in numeric mode while not characters. In **GSEAtraining**, there is a `add_more_columns()` which adds more columns to the `tb@result` table.

```{r}
tb2 = add_more_columns(tb)
head(tb2@result)
```


## Practice 

### {.tabset}

#### Practice 1

Use the DE gene list in `webgestalt_example_gene_list.rds` to perform ORA on
GO/KEGG/MSigDB gene sets.

```{r}
genes = readRDS(system.file("extdata", "webgestalt_example_gene_list.rds", 
	package = "GSEAtraining"))
```

