---
title: "Topic 5-01: Similarities of terms"
author: "Zuguang Gu <z.gu@dkfz.de>"
date: '`r Sys.Date()`'
output:
  html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Topic 5-01: Similarities of terms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


There are two types of methods to calculate similarities between functional terms.

- purely based on the number of member genes.
- based on the topology of the ontology.


We demonstrate it with significant terms from an ORA analysis.

```{r}
library(GSEAtraining)
lt = readRDS(system.file("extdata", "ora.rds", package = "GSEAtraining"))
diff_gene = lt$diff_gene
diff_gene = convert_to_entrez_id(diff_gene)
```

## Similarity based on gene counts

First we use a small collection of gene sets, the KEGG pathways.

```{r}
df = read.table(url("https://rest.kegg.jp/link/pathway/hsa"), sep = "\t")
df[, 1] = gsub("hsa:", "", df[, 1])
df[, 2] = gsub("path:",  "", df[, 2])
head(df)
```

We also retrieve the pathway names.

```{r}
pathway_names = read.table(url("https://rest.kegg.jp/list/pathway/hsa"), sep = "\t")
pathway_names = structure(gsub(" - Homo .*$", "", pathway_names[, 2]), names = pathway_names[, 1])
```

Perform ORA analysis with **clusterProfiler**. There are 33 significant pathways.

```{r}
library(clusterProfiler)
tb = enricher(diff_gene, TERM2GENE = df[, 2:1])
sig_pathways = as.data.frame(tb)$ID

gs_pathway = split(df[, 1], df[, 2])
gs_pathway = gs_pathway[sig_pathways]
```

There are four similarity methods which are based on the counts:

- Jaccard coeffcient
- Dice coeffcient
- Overlap coeffcient
- Kappa coeffcient

Since Dice coeffcient is just a simple transformation from the Jaccard coeffcient, we only
look at the following three methods:


```{r}
library(simplifyEnrichment)
m1 = term_similarity(gs_pathway, method = "jaccard")
m2 = term_similarity(gs_pathway, method = "overlap")
m3 = term_similarity(gs_pathway, method = "kappa")
```

Values on diagonals of the three symmetric matrices are 1. Since other values in the matrices
are much smaller than 1, we set the diagonals to NA to improve the heatmap visualization.

```{r, fig.width = 10, fig.height = 7}
diag(m1) = NA
diag(m2) = NA
diag(m3) = NA

library(ComplexHeatmap)
Heatmap(m1, name = "jaccard") + 
    rowAnnotation(pathway_name = anno_text(pathway_names[rownames(m1)]))
Heatmap(m2, name = "overlap") + 
    rowAnnotation(pathway_name = anno_text(pathway_names[rownames(m1)]))
Heatmap(m3, name = "kappa") + 
    rowAnnotation(pathway_name = anno_text(pathway_names[rownames(m1)]))
```

In previous examples, we use all genes in the gene sets. Next we restrict only
to the DE genes.

```{r, fig.width = 10, fig.height = 7}
m1 = term_similarity(gs_pathway, method = "jaccard", all = diff_gene)
m2 = term_similarity(gs_pathway, method = "overlap", all = diff_gene)
m3 = term_similarity(gs_pathway, method = "kappa", all = diff_gene)

diag(m1) = NA
diag(m2) = NA
diag(m3) = NA

Heatmap(m1, name = "jaccard") + 
    rowAnnotation(pathway_name = anno_text(pathway_names[rownames(m1)]))
Heatmap(m2, name = "overlap") + 
    rowAnnotation(pathway_name = anno_text(pathway_names[rownames(m1)]))
Heatmap(m3, name = "kappa") + 
    rowAnnotation(pathway_name = anno_text(pathway_names[rownames(m1)]))
```

Next we use a much larger gene set collection: the GO gene sets:

```{r}
library(org.Hs.eg.db)
tb = enrichGO(gene = diff_gene, ont = "BP", OrgDb = org.Hs.eg.db)

sig_go = as.data.frame(tb)$ID
```

There are > 800 significant GO BP terms.

```{r}
gs_bp = get_GO_gene_sets_from_orgdb(org.Hs.eg.db)
gs_bp = gs_bp[sig_go]
```

```{r}
m1 = term_similarity(gs_bp, method = "jaccard")
m2 = term_similarity(gs_bp, method = "overlap")
m3 = term_similarity(gs_bp, method = "kappa")

diag(m1) = NA
diag(m2) = NA
diag(m3) = NA

Heatmap(m1, name = "jaccard", show_row_names = FALSE, show_column_names = FALSE,
    show_row_dend = FALSE, show_column_dend = FALSE)
Heatmap(m2, name = "overlap", show_row_names = FALSE, show_column_names = FALSE,
    show_row_dend = FALSE, show_column_dend = FALSE)
Heatmap(m3, name = "kappa", show_row_names = FALSE, show_column_names = FALSE,
    show_row_dend = FALSE, show_column_dend = FALSE)
```

Or only restrict in DE genes

```{r}
m1 = term_similarity(gs_bp, method = "jaccard", all = diff_gene)
m2 = term_similarity(gs_bp, method = "overlap", all = diff_gene)
m3 = term_similarity(gs_bp, method = "kappa", all = diff_gene)

diag(m1) = NA
diag(m2) = NA
diag(m3) = NA

Heatmap(m1, name = "jaccard", show_row_names = FALSE, show_column_names = FALSE,
    show_row_dend = FALSE, show_column_dend = FALSE)
Heatmap(m2, name = "overlap", show_row_names = FALSE, show_column_names = FALSE,
    show_row_dend = FALSE, show_column_dend = FALSE)
Heatmap(m3, name = "kappa", show_row_names = FALSE, show_column_names = FALSE,
    show_row_dend = FALSE, show_column_dend = FALSE)
```

## Semantic similarity based on ontology structures

There are also two major types of methods for semantic similarities.

- based on external gene annotations
- based on internal topology structure

### Based on gene annotation


```{r, echo = FALSE}
library(simona)
simona_opt$verbose = FALSE
```

```{r}
dag = create_ontology_DAG_from_GO_db("BP", org_db = org.Hs.eg.db)
```

The _Sim_Lin_1998_ is one of the classic semantic similarity method.

```{r}
m = term_sim(dag, sig_go, method = "Sim_Lin_1998")
diag(m) = NA
Heatmap(m, name = "Sim_Lin_1998", show_row_names = FALSE, show_column_names = FALSE,
    show_row_dend = FALSE, show_column_dend = FALSE)
```

### Purely based on the topology

The _Sim_WP_1994__ is one of the classic semantic similarity method.


```{r}
m = term_sim(dag, sig_go, method = "Sim_WP_1994")
diag(m) = NA
Heatmap(m, name = "Sim_WP_1994", show_row_names = FALSE, show_column_names = FALSE,
    show_row_dend = FALSE, show_column_dend = FALSE)
```

### Visualize the ontology

```{r, fig.width = 10, fig.height = 7}
dag_circular_viz(dag, highlight = sig_go)
```
