---
title: "day2_exercise"
output: html_document
---


Use the tiny GSEA framework on your own dataset and compare the results, e.g. use different gene-level methods and different set-level methods. You can use only one gene set or use all gene set in a database (e.g. all GO BP gene sets). 

			
gene-level	Gene-level transform        set-level	    name
tvalue	    abs                             mean	    tvalue-abs-mean
sam	        square	                        sum	        sam-square-sum
ttest    	binary	                        sum	        ttest-binary-sum
ttest	    binary	                        chisq	    ttest-binary-chisq
sam	        none	                        ks      	sam-ks
tvalue	    none	                        wilcox      tvalue-wilcox

 code is at https://gist.github.com/jokergoo/e8fff4a57ec59efc694b9e730da22b9f

```{r}
source("https://gist.githubusercontent.com/jokergoo/e8fff4a57ec59efc694b9e730da22b9f/raw/530d6d9eabb6a7c7c6b287549b8fb44efabccc5b/tiny_GSEA_framework.R")
```


```{r}
library(CePa)
condition = read.cls("P53.cls", treatment = "MUT", control = "WT")$label
expr = read.gct("P53_collapsed_symbols.gct")

condition = factor(condition)
```

Load Hallmark gene sets from MSigDB

```{r}
library(msigdbr)
gs = msigdbr(category = "H")
gs = split(gs$gene_symbol, gs$gs_name)
```

```{r}
tb = gsea_tiny(expr, condition, geneset = gs, 
    gene_level_method = "tvalue",
    transform = "abs",
    set_level_method = "mean")
```


```{r}
methods = data.frame(gene_level = c("tvalue", "sam", "ttest", "ttest", "sam", "tvalue"),
                     transform  = c("abs", "square", "binary", "binary", "none", "none"),
                     set_level  = c("mean", "sum", "sum", "chisq", "ks", "wilcox"))
rownames(methods) = c("tvalue-abs-mean", "sam-square-sum", "ttest-binary-sum",
                      "ttest-binary-chisq", "sam-ks", "tvalue-wilcox")
methods
```


```{r}
n_methods = nrow(methods)

res_list = list()

for(i in 1:n_methods) {
    cat("running method:", rownames(methods)[i], "\n")
    if(methods$transform[i] == "binary") {
        
        tb = gsea_tiny(expr, condition, geneset = gs, 
                gene_level_method = methods$gene_level[i],
                transform = methods$transform[i],
                binarize = function(x) ifelse(x < 0.05, 1, 0),
                set_level_method = methods$set_level[i])
        
    } else {
        
        tb = gsea_tiny(expr, condition, geneset = gs, 
                gene_level_method = methods$gene_level[i],
                transform = methods$transform[i],
                set_level_method = methods$set_level[i])
        
    }
    res_list[[i]] = tb
}
names(res_list) = rownames(methods)
```

res_list is a list of data frames

```{r}
library(ComplexHeatmap)

sig_list = lapply(res_list, function(tb) {
    rownames(tb)[tb$p.value < 0.05]
})
cm = make_comb_mat(sig_list)

UpSet(cm,
    top_annotation = upset_top_annotation(cm, add_numbers = TRUE),
    right_annotation = upset_right_annotation(cm, add_numbers = TRUE)
)
```

```{r}
res_list_by_gene = list()

for(i in 1:n_methods) {
    cat("running method:", rownames(methods)[i], "\n")
    if(methods$transform[i] == "binary") {
        tb = gsea_tiny(expr, condition, geneset = gs, 
                gene_level_method = methods$gene_level[i],
                transform = methods$transform[i],
                binarize = function(x) ifelse(x < 0.05, 1, 0),
                set_level_method = methods$set_level[i], perm_type = "gene")
    } else {
        tb = gsea_tiny(expr, condition, geneset = gs, 
                gene_level_method = methods$gene_level[i],
                transform = methods$transform[i],
                set_level_method = methods$set_level[i], perm_type = "gene")
    }
    res_list_by_gene[[i]] = tb
}
names(res_list_by_gene) = rownames(methods)
```

```{r}
sig_list = lapply(res_list_by_gene, function(tb) {
    rownames(tb)[tb$p.value < 0.05]
})
cm = make_comb_mat(sig_list)

UpSet(cm,
    top_annotation = upset_top_annotation(cm, add_numbers = TRUE),
    right_annotation = upset_right_annotation(cm, add_numbers = TRUE)
)
```