---
title: "day3_piano"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE)
```

**piano** support many gene set-level statistics. It accepts input as matrix or 
as a vector of gene-level scores. In this example, we use the pre-computed gene-level
scores which is the t-value in a two-condition comparison:

```{r}
library(CePa)
condition = read.cls("P53.cls", treatment = "MUT", control = "WT")$label
expr = read.gct("P53_collapsed_symbols.gct")

library(genefilter)
s = rowttests(expr, factor(condition))[, "statistic"]
names(s) = rownames(expr)
```

We use the hallmark gene sets from MSigDB.

```{r}
library(msigdbr)
gs = msigdbr(category = "H")
```

**piano** needs the gene sets data to be stored in a object in a specific format:

```{r}
library(piano)
gs = loadGSC(gs[, c("gene_symbol", "gs_name")])
```

`runGSA()` runs GSEA analysis. The default gene-set-level method is "mean" (take the mean of gene-level statistics).

Note since the input is a vector of gene-level scores, p-values are calculated based on gene permutation.

```{r}
res = runGSA(geneLevelStats = s, gsc = gs)
```

`exploreGSAres()` visualizes the results via a Shiny app.

The result table can be extracted with `GSAsummaryTable()`:

```{r}
res_tb = GSAsummaryTable(res)
head(res_tb)
```

`runGSA()` can only run a single method. Here we use `lapply()` to run multiple methods:

```{r, message=FALSE, eval = !file.exists("piano_res_list.RData")}
all_set_stats = c("mean", "median", "fgsea", "page")
res_list = lapply(all_set_stats, function(stat) {
    runGSA(geneLevelStats = s, gsc = gs, geneSetStat = stat)
})
names(res_list) = all_set_stats
```

```{r, echo = FALSE, eval = !file.exists("piano_res_list.RData")}
save(res_list, file = "piano_res_list.RData")
```


```{r, echo = FALSE}
load("piano_res_list.RData")
```

There is a `consensusHeatmap()` function in **piano** which draws a heatmap of the top gene sets by averaging ranks from multiple methods.

```{r, fig.height=8}
consensusHeatmap(res_list, method = "mean")
```

Next we extract the p-values from all methods:

```{r}
p_list = lapply(res_list, function(x) {
    GSAsummaryTable(x)[, 3]
})
```


If we use **ComplexHeatmap** for heatmaps.

```{r}
library(ComplexHeatmap)
rank_mat = do.call(cbind, lapply(p_list, rank))
Heatmap(rank_mat, name = "p-value rank")
```



