---
title: "day3_goseq"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---

Only a significant diff gene list is needed for GOseq analysis.

```{r}
tb = read.table("E-GEOD-101794-analytics.tsv", sep = "\t", skip = 1)
tb = tb[, c(1, 3)]
colnames(tb) = c("gene", "p.value")
tb = tb[!is.na(tb$p.value), ]
tb$fdr = p.adjust(tb$p.value, "BH")
```

Construct a binary vector where 1 means the gene is diff.

```{r}
genes = ifelse(tb$fdr < 0.05, 1, 0)
genes[is.na(genes)] = 0
names(genes) = tb$gene
table(genes)
```

We estimate the distribution of p_diff against gene length:

```{r, message = FALSE}
library(goseq)
pwf = nullp(genes, "hg19", "ensGene")
```

Then use `goseq()` function to perform the test:

```{r, message = FALSE}
tb1 = goseq(pwf, "hg19", "ensGene")
head(tb1)
```

This time we keep the genes with p-values as `NA` (they are basically genes not expressed).

```{r}
tb = read.table("E-GEOD-101794-analytics.tsv", sep = "\t", skip = 1)
tb = tb[, c(1, 3)]
colnames(tb) = c("gene", "p.value")
tb$fdr = p.adjust(tb$p.value, "BH")
genes = ifelse(tb$fdr < 0.05, 1, 0)
genes[is.na(genes)] = 0
names(genes) = tb$gene
table(genes)
pwf = nullp(genes, "hg19", "ensGene")
```

This becomes interesting. If include non-expressed genes, the bias becomes visible... Note we assume non-expressed
genes are not differentially expressed.

So actually it means the bias comes from the p_expressed to gene length? This has to be studies 
with more details.

```{r, message = FALSE}
tb2 = goseq(pwf, "hg19", "ensGene")
```


We compare `tb1` and `tb2`:

```{r}
library(eulerr)
lt = list(
    test1 = tb1$category[p.adjust(tb1$over_represented_pvalue, "BH") < 0.05], 
    test2 = tb2$category[p.adjust(tb2$over_represented_pvalue, "BH") < 0.05]
)
plot(euler(lt), quantities = TRUE)
```

Does it mean the correction has no effect?

We also compare to ORA.

```{r}
tb = read.table("E-GEOD-101794-analytics.tsv", sep = "\t", skip = 1)
tb = tb[, c(2, 3)]
colnames(tb) = c("gene", "p.value")
tb$fdr = p.adjust(tb$p.value, "BH")
l = tb$fdr < 0.05; l[is.na(l)] = FALSE
diff_gene = tb$gene[l]
```

`diff_gene` are gene symbols. We convert them to Entrez IDs:

```{r}
convert_to_entrez_id = function(x) {
	if(is.matrix(x)) {
		map = cola:::guess_id_mapping(rownames(x))
		if(is.function(map)) {
		    new_rn = map(rownames(x))
		} else {
		    new_rn = map[rownames(x)]
		}
		l = is.na(new_rn)

		x = x[!l, , drop = FALSE]
		new_rn = new_rn[!l]

		x2 = do.call(rbind, tapply(1:nrow(x), new_rn, function(ind) {
			colMeans(x[ind, , drop = FALSE])
		}))
		return(x2)

	} else if(is.numeric(x)) {
		map = cola:::guess_id_mapping(names(x))
		x2 = s
		if(is.function(map)) {
		    names(x2) = map(names(x))
		} else {
		    names(x2) = map[names(x)]
		}
		x2 = x2[!is.na(names(x2))]
		x2 = tapply(x2, names(x2), mean)
		return(x2)
	} else {
	    map = cola:::guess_id_mapping(x)
	    if(is.function(map)) {
	        x2 = map(x)
	    } else {
	        x2 = map[x]
	    }
	    x2 = x2[!is.na(x2)]
	    x2 = unique(x2)
	    return(x2)
	}
}
diff_gene = convert_to_entrez_id(diff_gene)
```

We use **clusterProfiler** to perform ORA analysis:


```{r, message = FALSE}
library(clusterProfiler)
library(org.Hs.eg.db)
tb3 = enrichGO(gene = diff_gene, ont = "BP", OrgDb = org.Hs.eg.db)
```

Now we compare ORA to GOseq results. Note `tb1` and `tb2` contains results for all three ontologies (BP, CC, MF). Here
we only need BP.

```{r}
tb1 = tb1[tb1$ontology == "BP", ]
tb2 = tb2[tb2$ontology == "BP", ]

library(eulerr)
lt = list(
    test1 = tb1$category[p.adjust(tb1$over_represented_pvalue, "BH") < 0.05], 
    test2 = tb2$category[p.adjust(tb2$over_represented_pvalue, "BH") < 0.05],
    ora = tb3$ID[p.adjust(tb3$pvalue, "BH") < 0.05]
)
plot(euler(lt), quantities = TRUE)
```

Please note the difference may also come from the different implementation of the tools (e.g. GO versions, precessings), while not the methods selves.
