---
title: "day4_visualization"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, fig.height = 8)
```

We first demonstrate the visualization on ORA analysis.

```{r}
load("demo_ora.RData")
```

The following function converts gene symbols to Entrez IDs.

```{r}
convert_to_entrez_id = function(x, org_db = "org.Hs.eg.db") {
    if(is.matrix(x)) {
        map = cola:::guess_id_mapping(rownames(x), org_db = org_db)
        if(is.function(map)) {
            new_rn = map(rownames(x))
        } else {
            new_rn = map[rownames(x)]
        }
        l = is.na(new_rn)

        x = x[!l, , drop = FALSE]
        new_rn = new_rn[!l]

        x2 = do.call(rbind, tapply(1:nrow(x), new_rn, function(ind) {
            colMeans(x[ind, , drop = FALSE])
        }))
        return(x2)

    } else if(is.numeric(x)) {
        map = cola:::guess_id_mapping(names(x), org_db = org_db)
        x2 = x
        if(is.function(map)) {
            names(x2) = map(names(x))
        } else {
            names(x2) = map[names(x)]
        }
        x2 = x2[!is.na(names(x2))]
        x2 = tapply(x2, names(x2), mean)
        return(x2)
    } else {
        map = cola:::guess_id_mapping(x, org_db = org_db)
        if(is.function(map)) {
            x2 = map(x)
        } else {
            x2 = map[x]
        }
        x2 = x2[!is.na(x2)]
        x2 = unique(x2)
        return(x2)
    }
}
diff_gene = convert_to_entrez_id(diff_gene)
```

We perform ORA analysis with **clusterProfiler**. Note the function `enrichGO()` returns a `enrichResult` object
and its slot ("`result`) contains the result table. Here we also extract the result table for future use.

Also note by default `enrichGO()` only returns significant results, so here we set `pvalueCutoff = 1` to let
`enrichGO()` to return all results (both significant and insignificant).

```{r}
library(clusterProfiler)
library(org.Hs.eg.db)
res = enrichGO(gene = diff_gene, ont = "BP", OrgDb = org.Hs.eg.db, pvalueCutoff = 1)
tb = res@result
```

How many signfiicant GO terms (FDR < 0.05)?

```{r}
sum(tb$p.adjust < 0.05)
```

First we make barplot for the significant terms:

```{r}
library(enrichplot)
barplot(res, showCategory = 20)
barplot(res, showCategory = 20, color = "p.adjust")
```


In the result table, there are fractions of geneset gene in the diff gene list and fractions of geneset genes in the background:

```{r}
head(tb)
```

We can calculate a "log2 fold change" which measures the relative value of fractions of geneset genes in the diff list and in the background.

The fraction is represented as strings "10/20", so we need to parse them into R expression:

```{r}
gene_ratio = sapply(tb$GeneRatio, function(x) eval(parse(text = x)))
bg_ratio = sapply(tb$BgRatio, function(x) eval(parse(text = x)))
log2enrichment = log2(gene_ratio/bg_ratio)
```

It might be interesting to see the distribution of log2 enrichment.

```{r}
hist(log2enrichment)
```

For the ease of use, `log2enrichment` is appended to both `res@result` and `tb`:

```{r}
res@result$log2enrichment = log2enrichment
tb$log2enrichment = log2enrichment
```

Now we can set `log2enrichment` as values on x-axis in the barplot:

```{r}
barplot(res, showCategory = 20, x = "log2enrichment")
```

Making dot plot is almost the same as barplot. In dot plot, gene sets are by default reordered by values on x-axis.

Note x-axis starts from the minimal on x-axis.

```{r}
dotplot(res, showCategory = 20)
dotplot(res, showCategory = 20, x = "log2enrichment")
```

Network visualization. Note by default it only shows top 5 gene sets.

```{r}
cnetplot(res)
```


Heatmap-like plot to visualize gene-geneset relations:

```{r, fig.width=10}
heatplot(res, showCategory = 10)
```

Tree-like visualization. We first need to calculate pairwise similarity between GO terms (by default Jaccard coefficient).

```{r}
res = pairwise_termsim(res)
treeplot(res)
```

Enrichment map visualization.

```{r, fig.width=10}
emapplot(res)
```

UpSet plot.

```{r, fig.width = 16}
upsetplot(res)
```


Volcano plot:

```{r}
plot(tb$log2enrichment, -log10(tb$p.adjust))
abline(h = 3, lty = 2)  # FDR < 0.001
abline(v = 1, lty = 2)   # log2enrichment > 1
```

-------------------------------

Next we demonstrate visualization on GSEA results. First we perform GSEA also by **clusterProfiler**.
Note here we also set `pvalueCutoff = 1` because we need the full set of results for some plots.

```{r}
load("gene_diff_score.RData")
gene_diff_score = sort(gene_diff_score, decreasing = TRUE)
res_gsea = gseGO(geneList=gene_diff_score, OrgDb = org.Hs.eg.db, ont = "BP", pvalueCutoff = 1)
```

Let's check what columns are there in the result table:

```{r}
head(res_gsea)
```


In barplot, `NES` can be used for x-axis:

```{r}
enrichplot:::barplot.enrichResult(res_gsea, showCategory = 20, x = "NES")
dotplot(res_gsea, showCategory = 20, x = "NES")
```

Also, in Upset plot, now the distribution of gene-level statistics is added to each combination set.

```{r, fig.width = 12}
upsetplot(res_gsea)
```

Ridge plot visualizes the general distribution of gene-level statistics in each set.

```{r}
ridgeplot(res_gsea)
```

And the famous GSEA plot.

```{r}
gseaplot(res_gsea, geneSetID = 1)
```

```{r}
gseaplot2(res_gsea, geneSetID = 1)
```

We can also compare multiple gene set in one GSEA plot. In the following plot, we 
visualize the most up-regulated gene set and the most down-regulated gene set.

```{r}
ind = c(which.max(res_gsea@result$NES), which.min(res_gsea@result$NES))
gseaplot2(res_gsea, geneSetID = ind)
```


Note to make the vocalno plot. NES scores are used for x-axis:

```{r}
tb_gsea = res_gsea@result
plot(tb_gsea$NES, -log10(tb_gsea$p.adjust))
abline(h = -log10(0.05), lty = 2)  # FDR < 0.05
abline(v = c(-1, 1), lty = 2)   # NES > 1, NES < -1
```


